--- 
title: "Time Series Analysis and Forecasting with ADAM"
author: "Ivan Svetunkov"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
bibliography: [library.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: config-i1/adam
description: "This textbook explains how to do time series analysis and forecasting using Advanced Dynamic Adaptive Model, implemented in smooth package for R."
---

# Preface {-}

This textbook uses two packages from R, namely `greybox`, which focuses on forecasting using regression models, and `smooth`, which implements Single Source of Error (SSOE) state space models for purposes of time series analysis and forecasting. The textbook focuses on explaining how ADAM ("ADAM is Dynamic Adaptive Model" - recursive acronym), one of the `smooth` functions (introduced in v3.0.0) works, also showing how it can be used in practice with examples from R. ADAM is a state space model based on exponential smoothing in ETS format and ARIMA. It encompasses both models and is expanded by introducing:

1. Explanatory variables (including time varying parameters);
2. Multiple frequencies;
3. Handling of intermittent data (data with natural zeroes);
4. Handling of missing data;
5. Variables and components selection and combination;
6. Analysis of parameters of the model;
7. And other minor features.

All these extentions are needed in order to solve specific real life problems, so we will have examples and case studies later in the book, in order to see how all of this can be used.

If you want to run exampels from the textbook, two packages are needed:
```{r eval=FALSE}
install.packages("greybox")
install.packages("smooth")
```

Some explanations of functions from the packages are given in my blog: [Package greybox for R](https://forecasting.svetunkov.ru/en/category/r-en/greybox/), [Package smooth for R](https://forecasting.svetunkov.ru/en/category/r-en/smooth/).

A very important thing to note is that this textbook **does not use `tidyverse` packages**. I like base R, and, to be honest, I am sure that `tidyverse` packages are great, but I have never needed them in my research. So, I will not use pipeline operators, `tibble` or `tsibble` objects and `ggplot2`. It is assumed throughout the textbook that you can do all those nice tricks on your own if you want to.

If you want to get in touch with me, there are lots of ways to do that: comments section on any page of [my website](https://forecasting.svetunkov.ru/), [my Russian website](https://ivan.svetunkov.ru/), [vk.com](https://vk.com/isvetunkov), [Facebook](https://www.facebook.com/isvetunkov), [Linkedin](https://www.linkedin.com/in/isvetunkov), [Twitter](https://twitter.com/iSvetunkov).

You can also find me on [ResearchGate](https://www.researchgate.net/profile/Ivan_Svetunkov), [StackExchange](http://stats.stackexchange.com/users/87585/ivan-svetunkov) and [StackOverflow](http://stackoverflow.com/users/5296618/ivan-svetunkov), although I’m not really active there. Finally, I also have [GitHub](https://github.com/config-i1) account.

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```
