--- 
title: "Time Series Analysis and Forecasting with ADAM"
author: "Ivan Svetunkov"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
bibliography: [library.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: config-i1/adam
description: "This textbook explains how to do time series analysis and forecasting using Advanced Dynamic Adaptive Model, implemented in smooth package for R."
---

# Preface {-}

This textbook uses two packages from R, namely `greybox`, which focuses on forecasting using regression models, and `smooth`, which implements Single Source of Error (SSOE) state space models for purposes of time series analysis and forecasting. The textbook focuses on explaining how ADAM ("ADAM is Dynamic Adaptive Model" - recursive acronym), one of the `smooth` functions (introduced in v3.0.0) works, also showing how it can be used in practice with examples from R. So, if you want to run exampels from the textbook, two packages need to be installed:
```{r eval=FALSE}
install.packages("greybox")
install.packages("smooth")
```

Some explanations of functions from the packages are given in my blog: [Package greybox for R](https://forecasting.svetunkov.ru/en/category/r-en/greybox/), [Package smooth for R](https://forecasting.svetunkov.ru/en/category/r-en/smooth/).

If you want to get in touch with me, there are lots of ways to do that: comments section on any page of [my website](https://forecasting.svetunkov.ru/), [my Russian website](https://ivan.svetunkov.ru/), [vk.com](https://vk.com/isvetunkov), [Facebook](https://www.facebook.com/isvetunkov), [Linkedin](https://www.linkedin.com/in/isvetunkov), [Twitter](https://twitter.com/iSvetunkov).

You can also find me on [ResearchGate](https://www.researchgate.net/profile/Ivan_Svetunkov), [StackExchange](http://stats.stackexchange.com/users/87585/ivan-svetunkov) and [StackOverflow](http://stackoverflow.com/users/5296618/ivan-svetunkov), although I’m not really active there. Finally, I also have [GitHub](https://github.com/config-i1) account.

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```
