# ADAM for Intermittent Demand {#ADAMIntermittent}
So far we have discussed data that has regular occurrence. This is a characteristic of a well established product that is sold every observation. For example, daily sales of bread in a supermarket would have this regularity. However, there are time series, where non-zero values do not happen on every observation. In the context of demand forecasting, this is called "**Intermittent demand**". The conventional example of such demand is monthly sales of jet engines: they will contain a lot of zeroes, when nobody buys the product and then all of a sudden several units, again followed by zeroes.

One of the simplest definitions of intermittent demand is that *it is the demand that happens at irregular frequency*. While at a first glance it might seem that it is an exotic problem, intermittent demand can be encountered in many areas, when the frequency of measurement is high enough. For example, daily sales of a specific type of tomatoes in a store might exhibit regular demand, but the same sales on hourly or minute frequency would exhibit intermittence. So, the problem is universal and might appear in almost any context.

Sometimes the term "count data" (or "integer-valued data") is used in a similar context, but there is a difference between this term and intermittent data. "Count data" implies that demand can take integer values only and can be typically modelled via Poisson, Binomial or Negative Binomial distributions. It does not necessarily contain zeroes and does not explicitly allow demand to happen at random. If there are zeroes, then it is assumed that they are just one of the possible values of a distribution. In case of intermittent demand, we explicitly acknowledge that demand might not happen, but if it happens then the demand size will be greater than zero. Furthermore, intermittent demand does not necessarily need to be integer-valued. For example, daily energy consumption for charging electric vehicles would typically be intermittent (because the vehicle owners do not charge them every day), but the non-zero consumption will not be integer. Having said that, count distributions can be used in some cases of intermittent demand, but they do not necessarily always provide a good approximation of complex reality.

Before we move towards the proper discussion of the topic in context of ADAM, we should acknowledge that at the heart of what follows, there lies the following model [@Croston1972]:
\begin{equation}
  y_t = o_t z_t ,
  (\#eq:IntermittentDemandModel)
\end{equation}
where $o_t$ is the demand occurrence variable, which can be either zero or one and has some probability of occurrence, $z_t$ is the demand sizes captured by a model (for example, ADAM ETS) and $y_t$ is the final observed demand. This model in context of intermittent demand was originally proposed by @Croston1972, but similar models (e.g. Hurdle and Zero Inflated Poisson) exist in other, non-forecasting related contexts.

In this chapter we will discuss the intermittent state space model that \@ref(eq:IntermittentDemandModel), both parts of which can be modelled via ADAM models, and we will see how they can be used, what they imply and how they connect to the convetional regular demand. If ETS model is used for $z_t$ then \@ref(eq:IntermittentDemandModel) is called iETS. So, iETS(M,N,N) model refers to the intermittent state space model, where demand sizes are modelled via ETS(M,N,N). ETS can also be used for occurrence part of the model, so if the discussion is focused on demand occurrence part of the model (as in Subsection \@ref(ADAMOccurrence)), we will use ``oETS'' instead.

While ARIMA can be used in this context as well, it is not yet implemented for the occurrence part of the model. So we will focus the discussio on ADAM ETS. Furthermore, depending on how the occurrence part is modelled, these notations can be expanded to include references to specific parts of the occurrence part of the model. This is discussed in detail in Subsection \@ref(ADAMOccurrence).

This chapter is based on @Svetunkov2019a.


## Occurrence part of the model {#ADAMOccurrence}
The general model \@ref(eq:IntermittentDemandModel) assumes that demand occurs randomly and that the variable $o_t$ can be either zero (no demand) or one (there is some demand). While this process can be modelled using different distributions, we propose using Bernoulli with a time varying probability (in the most general case):
\begin{equation}
    o_t \sim \text{Bernoulli} \left(p_t \right) ,
  (\#eq:OccurrenceModelBernoulli)
\end{equation}
where $p_t$ is the probability of occurrence. In this section we will discuss different types of models for this probability. Each one has some idea behind it, and there are different mechanisms of the model construction, estimation, error calculation, update of the states and the generation of forecasts for each of them.

The occurrence part of the model can be estimated via the likelihood, which in the most general case is:
\begin{equation}
	\ell \left(\boldsymbol{\theta} | o_t \right) = {\sum_{o_t=1}} \log(\hat{p}_t) + {\sum_{o_t=0}} \log(1-\hat{p}_t) ,
  (\#eq:oETSLikelihood)
\end{equation}
where $\hat{p}_t$ is the in-sample conditional one step ahead expectation of the probability on observation $t$, given the information on observation $t-1$, which depends on the vector of estimated parameters $\boldsymbol{\theta}$. The derivation of the formula \@ref(eq:oETSLikelihood) is discussed in more detail later in this chapter.

In order to demonstrate the difference between specific types of oETS models, we will use the following artificial data:
```{r artificialData}
y <- ts(c(rpois(20,0.25),rpois(20,0.5),rpois(20,1),rpois(20,2),rpois(20,3),rpois(20,5)))
```

Here how the data looks like:

```{r}
plot(y)
```

The probability of occurrence in this model increases together with the demand sizes. This example corresponds to the situation of intermittent demand of a product becoming regular.


### Fixed probability model, oETS$_F$
We start with the simplest case of the fixed probability of occurrence, oETS$_F$ model:
\begin{equation}
	o_t \sim \text{Bernoulli}(p) ,
	(\#eq:oETSFixed)
\end{equation}
This model assumes that demand happens with the same probability no matter what. This might sound exotic, because in practice, there might be many factors influencing customers' desire to purchase and the impact of these factors might change over the year. But this is a very basic model, which at least can be used as a benchmark on intermittent demand data. Furthermore, it might be suitable for modelling demand on expensive high-tech products, such as jet engines, which is ``very slow'' in its nature and typically does not evolve much over time.

When estimated via maximisation of likelihood function \@ref(eq:oETSLikelihood), the probability of occurrence is equal to:
\begin{equation}
	\hat{p} = \frac{T_1}{T},
	(\#eq:oETSFixedProbabilityMLE)
\end{equation}
where $T_1$ is the number of non-zero observations and $T$ is the number of all the available observations in sample.

The occurrence part of the model, oETS$_F$ can be constructed using `oes()` function from `smooth` package:

```{r oETSFExample1}
oETSFModel1 <- oes(y, occurrence="fixed", h=10, holdout=TRUE)
oETSFModel1
plot(oETSFModel1)
```

The plot above demonstrates the dynamics of the occurrence variable $o_t$ and the fitted and predicted probabilities. The oETS$_F$ model produces the straight line for the probability of `r round(oETSFModel1$initial,2)`, ignoring the fact that in our example the probability of occurrence has increased over time.


### Odds ratio model, oETS$_O$
In this model, it is assumed that the update of the probability is driven by the occurrence of variable. It is more complicated than the previous as the probability now changes over time and can be modelled, for example, with ETS(M,N,N) model:
\begin{equation}
    \begin{aligned}
		& o_t \sim \text{Bernoulli} \left(p_t \right) \\
		& p_t = \frac{\mu_{a,t}}{\mu_{a,t}+1} \\
		& a_t = l_{a,t-1} \left(1 + \epsilon_{a,t} \right) \\
		& l_{a,t} = l_{a,t-1}( 1  + \alpha_{a} \epsilon_{a,t}) \\
		& \mu_{a,t} = l_{a,t-1}
    \end{aligned},
	(\#eq:oETSOddsMNN)
\end{equation}
where $l_{a,t}$ is the unobserved level component, $\alpha_{a}$ is the smoothing parameter, the error term $1+\epsilon_{a,t}$ is positive, has means of one and follows an unknown distribution and $\mu_{a,t}$is the conditional expectation for the unobservable shape parameters $a_t$. The measurement and transition equations in \@ref(eq:oETSOddsMNN) can be substituted by any other ADAM ETS or ARIMA model if it is reasonable to assume that the dynamics of probability has some additional components. This model is called ``odds ratio'', because the probability of occurrence in \@ref(eq:oETSOddsMNN) is calculated using the classical logistic transform. This also means that $\mu_{a,t}$ is equal to:
\begin{equation} \label{eq:oETS_O_oddsRatio}
	\mu_{a,t} = \frac{p_t}{1 - p_t} .
\end{equation}
When $\mu_{a,t}$ increases in the oETS\textsubscript{O} model, the odds ratio increases as well, meaning that the probability of occurrence goes up. @Svetunkov2019a explain that this model is in theory more appropriate for the demand on products becoming obsolescent, but given the updating mechanism it should also work fine on other types of intermittent demand data.

When it comes to the application of the model to the data, its construction is done via the following set of equations (example with oETS$_O$(M,N,N)):
\begin{equation}
    \begin{aligned}
        & \hat{p}_t = \frac{\hat{\mu}_{a,t}}{\hat{\mu}_{a,t}+1} \\
		& \hat{\mu}_{a,t} = \hat{l}_{a,t-1} \\
		& \hat{l}_{a,t} = \hat{l}_{a,t-1}( 1  + \hat{\alpha}_{a} e_{a,t}) \\
		& 1+e_{a,t} = \frac{u_t}{1-u_t} \\
		& u_{t} = \frac{1 + o_t - \hat{p}_t}{2}
    \end{aligned},
	(\#eq:oETSOddsMNNEstimated)
\end{equation}
where $e_{a,t}$ is the proxy for the unobservable error term $\epsilon_{a,t}$ and $\hat{\mu}_t$ is the estimate of $\mu_{a,t}$, conditional expectation discussed in Subsection \@ref(ETSTaxonomyMaths). If a multiple steps ahead forecast for the probability is needed from this model, then the formulae discussed in Subsection \@ref(ETSTaxonomyMaths) can be used to get $\hat{\mu}_{a,t}$, which then can be inserted in the first equation of \@ref(eq:oETSOddsMNNEstimated) to get the final conditional multiple steps ahead probability of occurrence.

Finally, in order to estimate the model \@ref(eq:oETSOddsMNNEstimated), the likelihood \@ref(eq:oETSLikelihood) can be used.

The occurrence part of the model oETS$_O$ is constructed using the very same `oes()` function, but also allows specifying the ETS model to use. For example, here is the oETS$_O$(M,M,N) model:
```{r oETSOExample1}
oETSOModel <- oes(y, model="MMN", occurrence="odds-ratio", h=10, holdout=TRUE)
oETSOModel
plot(oETSOModel)
```

The constructed model introduces the trend component, capturing the changing probability of occurrence and reflecting well the fact that it increases over time.


### Inverse odds ratio model, oETS$_I$
Using similar approach to the oETS$_O$ model, we can formulate the "inverse odds ration" model oETS$_I$(M,N,N):
\begin{equation}
    \begin{aligned}
		& o_t \sim \text{Bernoulli} \left(p_t \right) \\
		& p_t = \frac{1}{1+\mu_{b,t}} \\
		& b_t = l_{b,t-1} \left(1 + \epsilon_{b,t} \right) \\
		& l_{b,t} = l_{b,t-1}( 1  + \alpha_{b} \epsilon_{b,t}) \\
		& \mu_{b,t} = l_{b,t-1}
    \end{aligned},
	(\#eq:oETSInverseOddsMNN)
\end{equation}
where similarly to \@ref(eq:oETSOddsMNNEstimated), $l_{b,t}$ is the unobserved level component, $\alpha_{b}$ is the smoothing parameter, the error term $1+\epsilon_{b,t}$ is positive, has means of one and follows an unknown distribution and $\mu_{b,t}$is the conditional expectation for the unobservable shape parameters $b_t$. The main difference of this model with the previous is in the different mechanism of probability calculation, which focuses on the probability of "inoccurrence", i.e. on zeroes of data rather than on ones. This type of model should be more appropriate for cases of demand building up [@Svetunkov2019a]. The probability calculation mechanism in \@ref(eq:oETSInverseOddsMNN) implies that $\mu_{b,t}$ is equal to:
\begin{equation}
	\mu_{b,t} = \frac{1-p_t}{p_t} .
\end{equation}

The construction of the model \@ref(eq:oETSInverseOddsMNN) is similar to \@ref(eq:oETSOddsMNNEstimated):
\begin{equation}
    \begin{aligned}
        & \hat{p}_t = \frac{1}{1+\hat{\mu}_{b,t}} \\
		& \hat{\mu}_{b,t} = \hat{l}_{b,t-1} \\
		& \hat{l}_{b,t} = l_{b,t-1}( 1  + \hat{\alpha}_{b} e_{b,t}) \\
		& 1+e_{b,t} = \frac{1-u_t}{u_t} \\
		& u_{t} = \frac{1 + o_t - \hat{p}_t}{2}
    \end{aligned},
	(\#eq:oETSInverseOddsMNNEstimated)
\end{equation}
where $e_{b,t}$ is the proxy for the unobservable error term $\epsilon_{b,t}$ and $\hat{\mu}_{b,t}$ is the estimate of $\mu_{b,t}$. Once again, we refer an interested reader to Subsection \@ref(ETSTaxonomyMaths) for the discussion of the multiple steps ahead conditional expectations from the model.

@Svetunkov2019a show that the oETS$_I$(M,N,N) model can also be estimated using Croston's method, as long as we can assume that the probability does not change over time substantially. In this case the demand intervals can be used instead of $\hat{\mu}_{b,t}$ in \@ref(eq:oETSInverseOddsMNNEstimated). So the iETS(M,N,N)$_I$(M,N,N) can be considered as the model underlying Croston's method.

The function `oes()` implements the oETS$_I$ model as well. For example, here is the oETS$_I$(M,M,N) model:
```{r oETSIExample1}
oETSIModel <- oes(y, model="MMN", occurrence="inverse-odds-ratio", h=10, holdout=TRUE)
oETSIModel
plot(oETSIModel)
```

Similarly to the oETS$_O$, the model captures the trend in the probability of occurrence, but will have different smoothing parameters.


### General oETS model, oETS$_G$
Uniting the models oETS$_O$ with oETS$_I$, we can obtain the "general" model, which in the most general case can be summarised in the following way:
\begin{equation}
    \begin{aligned}
		& p_t = f(\mu_{a,t}, \mu_{b,t}) \\
		& a_t = w_a(\mathbf{v}_{a,t-\mathbf{l}}) + r_a(\mathbf{v}_{a,t-\mathbf{l}}) \epsilon_{a,t} \\
		& \mathbf{v}_{a,t} = f_a(\mathbf{v}_{a,t-\mathbf{l}}) + g_a(\mathbf{v}_{a,t-\mathbf{l}}) \epsilon_{a,t} \\
		& b_t = w_a(\mathbf{v}_{b,t-\mathbf{l}}) + r_a(\mathbf{v}_{b,t-\mathbf{l}}) \epsilon_{b,t} \\
		& \mathbf{v}_{b,t} = f_a(\mathbf{v}_{b,t-\mathbf{l}}) + g_a(\mathbf{v}_{b,t-\mathbf{l}}) \epsilon_{b,t}
    \end{aligned} ,
  (\#eq:oETSG)
\end{equation}
where $\epsilon_{a,t}$, $\epsilon_{b,t}$, $\mu_{a,t}$ and $\mu_{b,t}$ have been defined in previous subsections. Note that two models similar to the one discussed in Section \@ref(ADAMETSIntroduction) are used for modelling of $a_t$ and $b_t$. The general formula for the probability in case of the multiplicative error model is:
\begin{equation}
    p_t = \frac{\mu_{a,t}}{\mu_{a,t}+\mu_{b,t}} ,
  (\#eq:oETSMZZ)
\end{equation}
while for the additive one, it is:
\begin{equation}
    p_t = \frac{\exp(\mu_{a,t})}{\exp(\mu_{a,t})+\exp(\mu_{b,t})} .
  (\#eq:oETSAZZ)
\end{equation}
This is because both $\mu_{a,t}$ and $\mu_{b,t}$ need to be strictly positive, while the additive error models support the real plane. The canonical oETS model assumes that the pure multiplicative model is used for the both $a_t$ and $b_t$. This type of model is positively defined for any values of error, trend and seasonality, which is essential for the values of $a_t$ and $b_t$ and their expectations. If a combination of additive and multiplicative error models is used, then the additive part should be exponentiated prior to the usage of the formulae for the calculation of the probability.

An example of the oETS model is oETS$_G$(M,N,N)(M,N,N):
\begin{equation}
    \begin{aligned}
		& o_t \sim \text{Bernoulli} \left(p_t \right) \\
		& p_t = \frac{\mu_{a,t}}{\mu_{a,t}+\mu_{b,t}} \\
		\\
		& a_t = l_{a,t-1} \left(1 + \epsilon_{a,t} \right) \\
		& l_{a,t} = l_{a,t-1}( 1  + \alpha_{a} \epsilon_{a,t}) \\
		& \mu_{a,t} = l_{a,t-1} \\
		\\
		& b_t = l_{b,t-1} \left(1 + \epsilon_{b,t} \right) \\
		& l_{b,t} = l_{b,t-1}( 1  + \alpha_{b} \epsilon_{b,t}) \\
		& \mu_{b,t} = l_{b,t-1} \\
    \end{aligned},
  (\#eq:oETSGExample)
\end{equation}
where all the parameters have already been defined in previous subsections. More advanced models can be constructing for $a_t$ and $b_t$ by specifying the ETS models for each part and / or adding explanatory variables. The construction of this model is done via the following set of equations:
\begin{equation}
	\begin{aligned}
		& e_{a,t} = \frac{u_t}{1-u_t} -1 \\
		& \hat{a}_t = \hat{l}_{a,t-1} \\
		& \hat{l}_{a,t} = \hat{l}_{a,t-1}( 1  + \alpha_{a} e_{a,t}) \\
		& e_{b,t} = \frac{1-u_t}{u_t} -1 \\
		& \hat{b}_t = \hat{l}_{b,t-1} \\
		& \hat{l}_{b,t} = \hat{l}_{b,t-1}( 1  + \alpha_{b} e_{b,t})
	\end{aligned} .
  (\#eq:oETSGExampleEstimated)
\end{equation}
<!-- The initialisation of the parameters of the oETS$_G$ model is done separately for the variables $a_t$ and $b_t$, based on the principles, described above for the oETS$_O$ and oETS$_I$. -->

There is a separate function for the oETS$_G$ model, called `oesg()`. It has twice more parameters than `oes()`, because it allows fine tuning of the models for the both variables $a_t$ and $b_t$. This gives an additional flexibility. For example, here is how we can use ETS(M,N,N) for the $a_t$ and ETS(A,A,N) for the $b_t$, resulting in oETS$_G$(M,N,N)(A,A,N):
```{r oETSGExample1}
oETSGModel1 <- oesg(y, modelA="MNN", modelB="AAN", h=10, holdout=TRUE)
oETSGModel1
plot(oETSGModel1)
```

We can also analyse separately models for $a_t$ and $b_t$. Here is, for example, the model A:
```{r}
oETSGModel1$modelA
```

The experiments that I have done so far show that oETS$_G$ very seldomly brings improvements in comparison with oETS$_O$ or oETS$_I$ in terms of forecasting accuracy. Besides, selecting models for each of the parts is a challenging task. So, this model is theoretically nice, being more general than the other oETS models, but is not very practical. Still it is useful because we can introduce different oETS model by restricting $a_t$ and $b_t$. For example, we can get:

1. oETS$_F$, when $\mu_{a,t} = \text{const}$, $\mu_{b,t} = \text{const}$ for all $t$;
2. oETS$_O$, when $\mu_{b,t} = 1$ for all $t$;
3. oETS$_I$, when $\mu_{a,t} = 1$ for all $t$;
4. oETS$_D$, when $\mu_{a,t} + \mu_{b,t} = 1$, $\mu_{a,t} \leq 1$ for all $t$ (discussed in the next subsection);
5. oETS$_G$, when there are no restrictions.


### Direct probability model, oETS$_D$
The last model in the family of oETS is the "Direct probability". It appears, when the following restriction is imposed on the oETS$_G$:
\begin{equation}
    \mu_{a,t} + \mu_{b,t} = 1, \mu_{a,t} \in [0, 1] .
  (\#eq:iETSGRestriction)
\end{equation}
This restriction is inspired by the mechanism for the probability update proposed by @Teunter2011 (TSB method). In their paper they use [SES](#SES) to model the time varying probability of occurrence. Based on this idea and the restriction \@ref(eq:iETSGRestriction) we can formulate oETS$_D$(M,N,N) model, which will underly the occurrence part of the TSB method:
\begin{equation}
    \begin{aligned}
		& o_t \sim \text{Bernoulli} \left(\mu_{a,t} \right) \\
		& a_t = l_{a,t-1} \left(1 + \epsilon_{a,t} \right) \\
		& l_{a,t} = l_{a,t-1}( 1  + \alpha_{a} \epsilon_{a,t}) \\
		& \mu_{a,t} = \min(l_{a,t-1}, 1)
    \end{aligned}.
  (\#eq:oETSD)
\end{equation}
There is also an option with the additive error for the occurrence part (also underlying TSB), which has a different, more complicated form:
\begin{equation}
    \begin{aligned}
		& o_t \sim \text{Bernoulli} \left(\mu_{a,t} \right) \\
		& a_t = l_{a,t-1} + \epsilon_{a,t} \\
		& l_{a,t} = l_{a,t-1}  + \alpha_{a} \epsilon_{a,t} \\
		& \mu_{a,t} = \max \left( \min(l_{a,t-1}, 1), 0 \right)
    \end{aligned}.
  (\#eq:oETSDAdditive)
\end{equation}

The estimation of the oETS$_D$(M,N,M) model can be done using the following set of equations:
\begin{equation}
	\begin{aligned}
		& \hat{\mu}_{a,t} = \hat{l}_{a,t-1} \\
		& \hat{l}_{a,t} = \hat{l}_{a,t-1}( 1  + \hat{\alpha}_{a} e_{a,t})
	\end{aligned},
  (\#eq:ISSETSMNNProbabilityEstimate)
\end{equation}
where
\begin{equation}
	e_{a,t} = \frac{o_t (1 - 2 \kappa) + \kappa - \hat{\mu}_{a,t}}{\hat{\mu}_{a,t}},
  (\#eq:oETSDMultiplicativeError)
\end{equation}
and $\kappa$ is a very small number (for example, $\kappa = 10^{-10}$), needed only in order to make the model estimable. The estimate of the error term in case of the additive model is much simpler and does not need any specific tricks to work:
\begin{equation}
	e_{a,t} = o_t - \hat{\mu}_{a,t} ,
  (\#eq:oETSDAdditiveError)
\end{equation}
which is directly related to TSB method. Note that equation \@ref(eq:ISSETSMNNProbabilityEstimate) does not contain $\min$ function, because the estimated error \@ref(eq:oETSDMultiplicativeError) will always guarantee that the level will lie between 0 and 1 as long as the smoothing parameter lies in the [0, 1] region (which is the conventional assumption for both ETS(A,N,N) and ETS(M,N,N) models). This also applies for the oETS$_D$(A,N,N) model, where the $\max$ and $\min$ functions can be dropped as long as the smoothing parameter lies in [0,1].

When it comes to initialising the oETS$_D$ model, the values are calculated directly from the data without any additional transformations.

Here's an example of the application of the oETS$_D$(M,M,N) to the same artificial data:
```{r oETSDExample1}
oETSDModel <- oes(y, model="MMN", occurrence="d", h=10, holdout=TRUE)
oETSDModel
plot(oETSDModel)
```

The empirical analysis I have done so far on different datasets shows that oETS$_D$ model works efficiently in many cases and produces accurate forecasts. So, if you are unsure, which of the oETS models to choose for your intermittent data, I would recommend starting with oETS$_D$.


## Demand sizes part of the model
So far we have discussed the occurrence part $o_t$ of the model and how to capture the probability of demand occurrence $p_t$. But this is only a half of the intermittent state space model. The second one is the model for the demand sizes $z_t$, which focuses on how many units of product will be sold if our customers decide to buy in a specific period of time. This can be modelled with any ADAM model, but has its own implications.

We start discussion with analysis of iETS(M,N,N)$_F$ model, which can be formulate as:
\begin{equation}
    \begin{aligned}
        & y_t = o_t z_t  \\
		& z_t = l_{z,t-1}(1 + \epsilon_{z,t}) \\
		& l_{z,t} = l_{z,t-1}(1  + \alpha_{z} \epsilon_{z,t}) \\
		& o_t \sim \text{Bernoulli}(p) \\
    \end{aligned},
	(\#eq:iETSMNNFixed)
\end{equation}
where the subscript $z$ refers to the components and parameters of demand sizes. This model assumes that there is always a potential demand on the product which evolves over time, even when $o_t=0$, we just do not always observe it. All the properties of this model have alread been discussed in chapter \@ref(ADAMETSPureMultiplicative). The main challenge appears, when this model needs to be constructed and estimated, because $z_t$ is not observable, when $o_t=0$. In these situations the error term cannot be esimated, but according to the model it still exists, thus impacting the level of demand $l_{z,t}$. In order to construct the model in the cases of no demand, we propose taking the conditional expectation for these periods, given the last available non-zero observations. This means that the model can be constructed using the following set of equations
\begin{equation}
    \begin{aligned}
		& e_{z,t} = \frac{z_t - \hat{\mu}_{z,t}}{\hat{\mu}_{z,t}}, \text{ when } o_t=1 \\
		& \hat{\mu}_{z,t} = \hat{l}_{z,t-1} \\
		& \hat{l}_{z,t} = 
		\left \lbrace \begin{aligned}
                  & \hat{l}_{z,t-1} (1 + \hat{\alpha}_z e_t ), & \text{ when } o_t=1 \\
                  & \hat{l}_{z,t-1} , & \text{ when } o_t=0
            \end{aligned} \right.
    \end{aligned}.
	(\#eq:iETSMNNFixedConstruction)
\end{equation}
This is only possible if $\mathrm{E}(1+\epsilon_{z,t})=1$, which is an important assumption for multiplicative error models, discussed in Section \@ref(ADAMETSMultiplicativeDistributions). If this is violated, then the formula for the calculation of the level in \@ref(eq:iETSMNNFixedConstruction) will become more complicated, involving the expectation of products of random variables.

In a similar way, we can construct more complicated models for the demand sizes. In a more [general case](#ADAMETSIntroduction) this can be written as:
\begin{equation}
  \begin{aligned}
		& e_{z,t} = \frac{z_t - \hat{\mu}_{z,t}}{\hat{\mu}_{z,t}}, \text{ when } o_t=1 \\
        & \hat{\mathbf{v}}_{t} = 
		    \left \lbrace \begin{aligned}
		        & f(\hat{\mathbf{v}}_{t-\mathbf{l}}) + g(\hat{\mathbf{v}}_{t-\mathbf{l}}) e_t, & \text{ when } o_t=1 \\
		        & f(\hat{\mathbf{v}}_{t-\mathbf{l}}) , & \text{ when } o_t=0
            \end{aligned} \right.
  \end{aligned},
  (\#eq:iETSADAMStateSpaceConstruction)
\end{equation}
where all the functions and vectors have been defined for the original ADAM ETS model \@ref(eq:ETSADAMStateSpace) in Section \@ref(#ADAMETSIntroduction).

## Additive vs multiplicative models for demand sizes
iETS supports any type of ETS model, including [pure additive](#ADAMETSPureAdditive), [pure multiplicative](#ADAMETSPureMultiplicative) and [mixed](#ADAMETSMixedModels) ones. But the selection of the appropriate model should be done based on the understanding of the problem. Typically, we expect demands to be non-negative: people want to buy our product, and usually the business does not want to buy from customers. In this case, we should use pure multiplicative models, as they will always produce meaningful results, as long as the assumption of positivity of $(1+\epsilon_{z,t})$ holds. This is important because the data would typically have low volume and the model might generate unreasonable (negative) pont and interval forecasts if a non-positive distribution is used (e.g. [Normal](#distributionsNormal)). Thus, it is important to use either [Inverse Gaussian, or Gamma, or Log Normal distribution](#ADAMETSMultiplicativeDistributions) for the error term of the demand sizes part of the model, when the volume of data is low and you expect the non-zero values to be strictly positive.

The main difficulty with pure multiplicative models arrises from the construction point of view - as discussed in Section \@ref(pureMultiplicativeExpectationAndVariance) the point forecasts of such model in general do not correspond to the conditional expectations (the only exclusion is the ETS(M,N,N) model). At the same time, the construction of the model for demand sizes assumes that the conditional expectations are equal to point forecasts, when demand is not observed. If this is violated then \@ref(eq:iETSADAMStateSpaceConstruction) is no longer the correct way to construct the model. This problem becomes especially important for the models with multiplicative trend, where the conditional expectation might differ from point forecasts substantially.

The pure additive models do not have the issue with the conditional expectation and thus can be constructed easily in case of intermittent demand. But as discussed earlier, they might violated the non-negativity assumption of the model. So, in practice they should be used with care.



<!-- ## Intermittent ETS -->
<!-- Note that estimating seasonality on intermittent data is challenging -->


<!-- In the notation of the model iETS(M,N,N)$_G$(M,N,N)(M,N,N), the first brackets describe the ETS model for the demand sizes, the underscore letter points out at the specific subtype of model (whic is discussed in the next subsections), the second brackets describe the ETS model, underlying the variable $a_t$ and the last ones stand for the model for the $b_t$. If only one variable is needed (either $a_t$ or $b_t$), then the redundant brackets are dropped, so that the notation simplifies, for example, to: iETS(M,N,N)$_O$(M,N,N). If the same type of model is used for both demand sizes and demand occurrence, then the second brackets can be dropped as well, simplifying this further to: iETS(M,N,N)$_G$. Furthermore, the notation without any brackets, such as iETS$_G$ stands for a general class of a specific subtype of iETS model (so any error / trend / seasonality). Also, given that iETS$_G$ is the most general model of all iETS models, the "$G$" can be dropped, when the properties are applicable to all subtypes. Finally, the "oETS" notation is used when the occurrence part of the model is discussed explicitly, skipping the demand sizes. -->

<!-- Depending on the restrictions on $a_t$ and $b_t$, there can be several types of iETS models: -->

<!-- 1. iETS$_F$: $\mu_{a,t} = \text{const}$, $\mu_{b,t} = \text{const}$ - the model with the fixed probability of occurrence; -->
<!-- 2. iETS$_O$: $\mu_{b,t} = 1$ - the "odds ratio" model, based on the logistic transform of the $o_t$ variable; -->
<!-- 3. iETS$_I$: $\mu_{a,t} = 1$ - the "inverse odds ratio" model, based on the inverse logistic transform of the $o_t$ variable; -->
<!-- 4. iETS$_D$: $\mu_{a,t} + \mu_{b,t} = 1$, $\mu_{a,t} \leq 1$ - the direct probability model, where the $p_t$ is calculated directly from the occurrence variable $o_t$; -->
<!-- 5. iETS$_G$: No restrictions - the model based on the evolution of both $\mu_{a,t}$ and $\mu_{b,t}$. -->



<!-- ### Model estimation -->
<!-- The concentrated likelihood function for the iETS model is: -->
<!-- \begin{equation} \label{eq:LogNormalConcentratedLogLikelihood} \tag{2} -->
<!-- 	\ell(\boldsymbol{\theta}, \hat{\sigma}_\epsilon^2 | \textbf{Y}) = - \frac{1}{2} \left( T \log(2 \pi e \hat{\sigma}_\epsilon^2) + T_0 \right) - {\sum_{o_t=1}} \log(z_t) + {\sum_{o_t=1}} \log(\hat{p}_t) + {\sum_{o_t=0}} \log(1-\hat{p}_t) , -->
<!-- \end{equation} -->
<!-- where $\textbf{Y}$ is the vector of all the in-sample observations, $\boldsymbol{\theta}$ is the vector of parameters to estimate (initial values and smoothing parameters), $T$ is the number of all observations, $T_0$ is the number of zero observations, $\hat{\sigma_\epsilon}^2 = \frac{1}{T} \sum_{o_t=1} \log^2 \left(1 + \epsilon_t \right)$ is the scale parameter of the one-step-ahead forecast error for the demand sizes and $\hat{p}_t$ is the estimated probability of a non-zero demand at time $t$. This likelihood is used for the estimation of all the special cases of the iETS$_G$ model. -->

<!-- ### Model selection in iETS -->

<!-- ### Conditional means and variance -->




<!-- ## Other related models -->
<!-- occurrence with provided values -->
<!-- occurrence estimated by alm() -->
<!-- Estimate initial seasonal on aggregate level -->
