# Model selection and combinations in ADAM {#ADAMSelection}
When it comes to time series analysis and to forecasting a specific time series, there are several ways to decide, which model to use, and there are several dimensions, in which a decision needs to be made:

1. Which of the models to use: ETS / ARIMA / ETS + ARIMA / Regression / ETSX / ARIMAX / ETSX+ARIMA?
2. What components of the ETS model to select?
3. What order of ARIMA model to select?
4. Which of the explanatory variables to use?
5. What distribution to use?
6. What occurrence model to use?
7. Should we select model or combine forecasts from different ones?
8. Do we need all models in the pool?
9. How should we do all the above?

In this chapter, we discuss all aspects, related to model selection and combinations in ADAM. We will start the discussion with principles based on [information criteria](#modelSelection), we will then move to more complicated topics, related to pooling and then we will finish with selection and combinations based on rolling origin.

Before we do that, we need to recall the [distributional assumptions](#distributions) in ADAM, which play an important role if the model is estimated via the [maximisation of likelihood function](#ADAMETSEstimationLikelihood). In this case an information criterion (IC) can be calculated and used for the selection of the most appropriate model. Based on this, we can fit several ADAM models with different distributions and then select the one that leads to the lowest IC. Here is the list of the supported distributions in ADAM:

- Normal;
- Laplace;
- S;
- Generalised Normal;
- Log Normal;
- Inverse Gaussian;
- Gamma.

The function `auto.adam()` implements this automatic selection of distribution based on IC for the provided vector of `distribution` by user. This selection procedure can be combined together with other selection techniques for different elements of ADAM model discussed in the following sections of the textbook.

Here is an example of selection of distribution for a specific model, ETS(M,M,N) on Box-Jenkins data using `auto.adam()`:
```{r}
adamModel <- auto.adam(BJsales, model="MMN", silent=FALSE, h=10, holdout=TRUE)
adamModel
```

In this case the function has applied one and the same model but with different distributions, estimated it using likelihood and selected the one that has the lowest AICc value.


## ADAM ETS components selection {#ETSSelection}
Having 30 ETS models to choose from, the task of selecting the most appropriate one becomes challenging. @Petropoulos2018a show that human experts can do this task successfully if they need to choose, which components to include in time series. However, when you face the problem of fitting ETS to thousands of time series, the judgmental selection becomes infeasible. Using some sort of automatic selection becomes critically important.

The basic idea underlying the components selection in ETS is based on [information criteria](#modelSelection): we define a pool of models, we fit those models and we select the one that has the lowest information criterion. Using this approach in ETS context was first proposed by @Hyndman2002. Based on this, we can construct some pool of models (e.g. based on our understanding of the problem) and then select the one that is the most appropriate to our data. `adam()` function in `smooth` package supports the following options for the pools:

1. Pool of [all 30 models](#ETSTaxonomy), `model="FFF"`;
2. Pool of [pure additive models](#ADAMETSPureAdditive), `model="XXX"`. As an option, "X" can also be used to tell function to only try additive component on the selected place. e.g. `model="MXM"` will tell function to only test ETS(M,N,M), ETS(M,A,M) and ETS(M,Ad,M) models;
3. Pool of [pure multiplicative models](#ADAMETSPureMultiplicative), `model="YYY"`. Similarly to (2), we can tell `adam()` to only consider multiplicative component in a specific place. e.g. `model="YNY"` will consider only ETS(M,N,N) and ETS(M,N,M);
4. Pool of pure models only, `model="PPP"` - this is a shortcut for doing (2) and (3) and then selecting the best between the two pools;
5. Manual pool of models, which can be provided as a vector of models, for example: `model=c("ANN","MNN","ANA","AAN")`.

In the cases explained above, `adam()` will try different models and select the most appropriate one from the predefined pool. There is a trade-off, when deciding which pool to use: if you provide the bigger one, it will take more time to find the appropriate one and there is a risk of overfitting the data; if you provide the smaller pool, then the optimal model might be outside of the pool, giving you the sub-optimal one.

Furthermore, in some situations you might not need to go through all 30 models, because, for example, the seasonal component is not needed in the data. Trying out all the models would be just a waste of time. So, in order to address this issue, I have developed a branch-and-bound algorithm for the selection of the most appropriate ETS model, which is triggered via `model="ZZZ"` (the same mechanism is used in `es()` function). The idea of the algorithm is to drop the components that do not improve the model. Here how it works:

1. Apply ETS(A,N,N) to the data, calculate an information criterion (IC);
2. Apply ETS(A,N,A) to the data, calculate IC. If it is lower than (1), then this means that there is some sort of seasonal component in the data, move to step (3). Otherwise go to (4);
3. If (2) is lower than (1), then apply ETS(M,N,M) model and calculate IC. If it is lower than it means that the data exhibits multiplicative seasonality. Go to (4);
4. Fit the model with the additive trend component and the seasonal component selected from previous steps, which can be either "N", "A" or "M", depending on the IC value. Calculate IC for the new model and compare it with the best IC so far. If it is lower, then there is some trend component in the data. If it is not, then the trend component is not needed.

Based on these 4 steps, we can kick off the unneeded components and reduce the pool of models to test. For example, if the algorithm shows that seasonality is not needed, but there is a trend, then we only have 10 models to test overall instead of 30: ETS(A,N,N), ETS(A,A,N), ETS(A,Ad,N), ETS(M,N,N), ETS(M,M,N), ETS(M,Md,N), ETS(A,M,N), ETS(A,Md,N), ETS(M,A,N), ETS(M,Ad,N). Also, in steps (2) and (3), if there is a trend in the data, then the model will have higher than needed smoothing parameter $\alpha$, but the seasonality will play an important role in reducing the value of IC. This is why the algorithm is in general efficient. It might not guarantee that the optimal model will be selected all the time, but it reduces the computational time.

The branch-and-bound algorithm can be combined with different types of models and is in fact is also supported in `model="XXX"` and `model="YYY`, where the pool of models for steps (1) - (4) is restricted by the pure models only.

Finally, while the branch-and-bound algorithm is quite efficient, it might end up providing a mixed model, which might not be very suitable for the data. So, it is recommended to think of the possible pool of models prior to applying it to the data. For example, in some cases you might realise that additive seasonality is not needed, and that the data can be either non-seasonal or with multiplicative seasonality. In this case, you can explore the `model="YZY"` option, aligning the error term with the seasonal component.

Here is an example with automatically selected ETS model using the branch-and-bound algorithm described above:
```{r}
adamETSModel <- adam(AirPassengers, model="ZZZ", silent=FALSE, h=12, holdout=TRUE)
adamETSModel
```

In this specific example, the optimal model will coincide with the one selected via `model="FFF"` and `model="ZXZ"`, although this does not necessarily is the case universally.


## ADAM ARIMA order selection {#ARIMASelection}
While ETS has 30 models to choose from, ARIMA has many more options. For example, selecting the non-seasonal ARIMA with / without constant restricting the orders with $p \leq 3$, $d \leq 2$ and $q \leq 3$ leads to the combination of $3 \times 2 \times 3 \times 2 = 36$ possible ARIMA models. If we increase the possible orders to 5 or even more, we will need to go through hundreds of models. Adding the seasonal part increases this number by an order of magnitude. This means that we cannot just test all possible ARIMA models and select the most appropriate one, we need to be smart in the selection proces.

@Hyndman2008Forecast developed an efficient mechanism of ARIMA order selection based on statistical tests (for stationarity and seasonality), reducing the number of models to test to reasonable ammount. @Svetunkov2019 developed an alternative mechanism, relying purely on information criteria, which works especially well on seasonal data, but potentially may lead to models that overfit the data (this is implemented in `auto.ssarima()` and `auto.msarima()` functions in `smooth` package). We also have the [Box-Jenkins approach](#BJApproach) for ARIMA orders selection, which relies on the analysis of [ACF](#ACF) and [PACF](#PACF), but we should not forget the [limitations of that approach](#BJApproachSummary). Finally, @Sagaert2021 proposed the [stepwise trace forward approach](#ETSXSelection), which relies on partial correlations and uses the information criteria to test model on each iteration. Building upon all of that, I have developed the following algorithm for order selection of ADAM ARIMA:

1. Determine the order of differences by fitting all possible combinations of ARIMA models with $P_j=0$ and $Q_j=0$ for all lags $j$. This includes trying the models with and without the constant term. The order $D_j$ is then determined via the model with the lowest IC;
2. Then iteratively, starting from the highest seasonal lag and moving to the lag of 1 do for every lag $m_j$:
a. Calculate ACF of residuals of the model;
b. Find the highest value of autocorrelation coefficient that corresponds to the multiple of the respective seasonal lag $m_j$;
c. Define, what should be the order of MA based on the lag of the autocorrelation coefficient on the previous step and include it in the ARIMA model;
d. Calculate IC, and if it is lower than for the previous best model, leave the new MA order;
e. Repeat (a) - (d) while there is an improvement in IC;
f. Do steps (a) - (e) for AR order, substituting ACF with PACF of the residuals of the best model;
g. Move to the next seasonal lag;
3. Try out several restricted ARIMA models of the order $q=d$ (this is based on (1) and the restrictions provide by the user). The motivation for this comes from the idea of relation between [ARIMA and ETS](#ARIMAandETS).

As you can see, this algorithm relies on the idea of Box-Jenkins methodology, but takes it with a pinch of salt, checking every time if the proposed order is improving the model or not. The motivation for doing MA orders before AR is based on the understanding of what [AR model](#AR) implies for forecasting. In a way, it is safer to have ARIMA(0,d,q) model than ARIMA(p,d,0), because the former is less prone to overfitting than the latter. Finally, the proposed algorithm is faster than the algorithm of @Svetunkov2019 and is more modest in the number of selected orders of the model.

In order to start the algorithm, you would need to provide a parameter `select=TRUE` in the `orders`. Here is an example with Box-Jenkins data:
```{r}
adamARIMAModel <- adam(BJsales, model="NNN", orders=list(ar=3,i=2,ma=3,select=TRUE),
                       silent=FALSE, h=10, holdout=TRUE)
adamARIMAModel
```

In this example, `orders=list(ar=3,i=2,ma=3,select=TRUE)` tells function that the maximum orders to check are $p\leq 3$, $d\leq 2$ $q\leq 3$.


## Explanatory variables selection {#ETSXSelection}
There are different approaches for [automatic variables selection](#modelSelection), but not all of them are efficient in the context of dynamic models. For example, backward stepwise might be either not feasible in case of small samples or may take too much time to converge to an optimal solution (it has polynomial computational time). This is because the ADAMX model needs to be refitted and reestimated over and over again using recursive relations based, for example, on the state space model \@ref(eq:ETSXADAMStateSpacePureAdditiveFull). The classical stepwise forward might also be too slow, because it has polynomial computational time. So, there need to be some simplifications, which will make variables selection in ADAMX doable in a reasonable time.

In order to make the mechanism doable in a limitted time, we rely on @Sagaert2021 approach of stepwise trace forward selection of variables. It is the approach that uses the partial correlations between variables in order to identify, which of the variables to include on each iteration, and has because of that linear computational time. Still, doing that in the proper ADAMX would take more time than needed, so one of the possibles solutions is to do variables selection in ADAMX in the following steps:

1. Estimate and fit the ETS model;
2. Extract the residuals of the ETS model;
3. Select the most suitable variables, explaining the residuals, based on an information criterion;
4. Estimate the ADAMX model with the selected explanatory variables.

The residuals in step (2) might vary from model to model, depending on the type of the error term and the selected distribution:

- Normal, Laplace, S, Generalised Normal or Asymmetric Laplace: $e_t$;
- Additive error and Log Normal, Inverse Gaussian or Gamma: $\left(1+\frac{e_t}{\hat{y}_t} \right)$;
- Multiplicative error and Log Normal, Inverse Gaussian or Gamma: $1+e_t$.

So, the extracted residuals should be formulated based on the distributional assumptions of each model.

In R, step (3) is done using the `stepwise()` function from `greybox` package, which supports all the distributions discussed in the previous chapters.

While the suggested approach has obvious limitations (e.g. smoothing parameters can be higher than needed, explaining the variability otherwise explained by variables), it is efficient in terms of computational time.

In order to see how it works, we use SeatBelt data:
```{r}
SeatbeltsData <- Seatbelts[,c("drivers","kms","PetrolPrice","law")]
```
We have already [had a look at this data earlier](#ETSXRExample), so we can move directly to the selection part:

```{r}
adamModelETSXMNMSelect <- adam(SeatbeltsData,"MNM",h=12,holdout=TRUE,regressors="select")
plot(forecast(adamModelETSXMNMSelect,h=12,interval="prediction"))
summary(adamModelETSXMNMSelect)
```

Based on the summary from the model, we can see that neither `kms`, nor `PetrolPrice` improve the model in terms of AICc. We could check them manually in order to see if the selection worked out well in our case (construct sink regression as a benchmark):
```{r}
adamModelETSXMNMSink <- adam(SeatbeltsData,"MNM",h=12,holdout=TRUE)
summary(adamModelETSXMNMSink)
```

We can see that the sink regression model has a higher AICc value than the model with the selected variables, which means that the latter is closer to the "true model". While `adamModelETSXMNMSelect` might not be the best possible model in terms of information criteria, it is still a reasonable one and allows making different decisions. For example, we can see from the summary of the model that the introduction of law has reduced the number of accidents with drivers by approximately 23.79%. However, this is an average effect, and the true one lies somewhere between -34.06% and -13.53% (with the 95% confidence).


<!-- ## Outlier detection mechanism {#OutliersDetection} -->


<!-- ## Occurrence part selection {#Occurrence selection} -->


<!-- ## Rolling origin selection {#ROSelection} -->


<!-- ## Pooling -->

