# Model diagnostics {#diagnostics}
In this chapter we investigate how ADAM can be diagnosed and improved. The majority of topics will build upon the typical model assumptions discussed in Section \@ref(assumptions). Some of the assumptions cannot be diagnosed properly, but for the others there are some existing and well established instruments. We will consider the following assumptions and discuss how to check whether they are violated or not:

1. [Model is correctly specified](#assumptionsCorrectModel):
a. No omitted variables;
b. No redundant variables;
c. The necessary transformation of the variables are applied;
d. No outliers in the model.
2. [Residuals are i.i.d.](#assumptionsResidualsAreIID):
a. They are not autocorrelated;
b. They are homoscedastic;
c. The expectation of residuals is zero, no matter what;
d. The residuals follow the specified distribution;
e. The distribution of residuals does not change over time.
3. [The explanatory variables are not correlated with anything but the response variable](#assumptionsXreg);

All the model diagnostics is aimed at spotting patterns in residuals. If there are some, then something is probably missing in the model. In this chapter we will discuss, which instruments can be used to diagnose different types of assumptions

In order to make this more actionable, we will consider a conventional regression model on `Seatbelts` data. This can be estimated equally well either with `adam()` from `smooth` or `alm()` from `greybox`. In general, I recommend using `alm()`, when no dynamic elements are present in the model, but for illustrative purposes we will do this with `adam()`:
```{r}
adamModelSeat01 <- adam(Seatbelts,"NNN",formula=drivers~PetrolPrice+kms)
plot(adamModelSeat01,7)
```

This model has several issues, and in this chapter we will discuss how to diagnose and fix them.


## Model specification: Omitted variables {#diagnosticsOmitted}
We start with one of the most important assumptions for models: model has not omitted important variables. In general this is difficult to diagnose, because typically it is not possible what is missing if we do not have it in front of us. The best thing one can do is a mental experiment, trying to comprise a list of all theoretically possible variables that would impact the variable of interest. If you manage to come up with such a list and realise that some of variables are missing, the next step would be to either collect the variables themselves or their proxies. One way or another, we would need to add the missing information in the model.

In some cases we might be able to diagnose this. For example, with our regression model from the previous section, we have a set of variables that are not included in the model. A simple thing to do would be to see if the residuals of our model are correlated with any of the omitted variables. We can either produce scatterplots or calculate measures of association to see if there is some relation in the residuals that is not explained by the existing structure. I will use `assoc()` and `spread()` functions from `greybox` for this:

```{r}
# Create a new matrix, removing the variables that are already in the model
SeatbeltsWithResiduals <- cbind(as.data.frame(residuals(adamModelSeat01)), Seatbelts[,-c(2,5,6)])
colnames(SeatbeltsWithResiduals)[1] <- "residuals"
# Spread plot
greybox::spread(SeatbeltsWithResiduals)
```

`spread()` function automatically detects the type of variable and produces scatterplot / `boxplot()` / `tableplot()` between them, making the final plot more readable. The plot above tells us that residuals are correlated with `DriversKilled`, `front`, `rear` and `law`, so some of these variables can be added to the model to improve it. `VanKilled` might have a weak relation with `drivers`, but judging by description does not make sense in the model (this is a part of the `drivers` variable). In our case, it is safe to add these variables, because they make sense in explaining the number of injured drivers. However, I would not add `DriversKilled` as it seems not to drive the number of deaths and injuries, but is just correlated with it for obvious reasons (`DriversKilled` is included in `drivers`). We can also calculate measures of association between variables:
```{r}
greybox::assoc(SeatbeltsWithResiduals)
```
Technically speaking, the output of this function tells us that all variables are correlated with residuals and can be considered in the model. I would still prefer not to add `DriversKilled` in the model for the reasons explained earlier. We can construct a new model in the following way:
```{r}
adamModelSeat02 <- adam(Seatbelts,"NNN",formula=drivers~PetrolPrice+kms+front+rear+law)
plot(adamModelSeat02,7)
```

How can we know that we have not omitted any important variables in our new model? Unfortunately, there is no good way of knowing that. In general, we should use judgment in order to decide whether anything else is needed or not. But given that we deal with time series, we can analyse residuals over time and see if there is any structure left:

```{r}
plot(adamModelSeat02,8)
```

This plot shows that the model has not captured seasonality and that there is stil some structure left in the residuals. In order to address this, we will add ETS(A,N,A) element to the model:
```{r}
adamModelSeat03 <- adam(Seatbelts,"ANA",formula=drivers~PetrolPrice+kms+front+rear+law)
par(mfcol=c(1,2))
plot(adamModelSeat03,7:8)
```

This is much better. There is no apparent missing structure in the data and no apparent omitted variables. We can now move to the next steps of diagnostics.


## Model specification: Redundant variables
While there are some ways of testing for omitted variables, the redundant ones are very difficult to diagnose. Yes, we could look at the [significance of variables](#hypothesisTesting) or compare models with and without some variables based on [information criteria](#informationCriteria), but even if our approaches say that a variable is not significant, this does not mean that it is not needed in the model. There can be many reasons, why a test would fail to reject H$_0$ and AIC would prefer a model without the variable under consideration. So, it comes to using judgment, trying to figure out whether a variable is needed in the model or not.

In the example with Seatbelt data, `DriversKilled` would be a redundant variable. Let's see what happens with the model in this case:

```{r}
adamModelSeat04 <- adam(Seatbelts,"NNN",formula=drivers~PetrolPrice+kms+front+rear+law+DriversKilled)
par(mfcol=c(1,2))
plot(adamModelSeat04,7:8)
```

The residuals from this model look adequate, with only issue being the first 45 observations lying below the zero line. The summary of this model is:
```{r}
summary(adamModelSeat04)
```
The uncertainty around the parameter `DriversKilled` is narrow, showing that the variable has a positive impact on the `drivers`. However the issue here is not statistical, but rather fundamental: we have included the variable that is a part of our response variable. It does not explain why drivers get injured and killed, it just reflects a specific part of this relation. So it explains part of the variance, which should have been explained by other variables (e.g. `kms` and `law`), making them statistically not significant. So, based on technical analysis we would be inclined to keep the variable, but based on our understanding of the problem we should not.

If we have redundant variables in the model, then the model might overfit the data, leading to narrower prediction intervals and biased forecasts. The parameters of such model are typically unbiased, but inefficient.


## Model specification: Transformations {#diagnosticsTransformations}
The question of appropriate transformations for variables in the model is challenging, because it is difficult to decide, what sort of transformation is needed, if needed at all. In many cases, this comes to selecting between additive linear model and a multiplicative one. This implies that we compare the model:
\begin{equation}
	y_t = a_0 + a_1 x_{1,t} + \dots + a_n x_{n,t} + \epsilon_t,
  (\#eq:additiveModel)
\end{equation}
and
\begin{equation}
	y_t = \exp\left(a_0 + a_1 x_{1,t} + \dots + a_n x_{n,t} + \epsilon_t\right) .
  (\#eq:multiplicativeModel)
\end{equation}
The latter model is equivalent to the so called "log-linear" model, but can also include logarithms of explanatory variables instead of the variables themselves.

There are different ways to diagnose the problem with wrong transformations, which sometimes help in detecting it. The first one is the actuals vs fitted plot:

```{r}
plot(adamModelSeat03,1)
```

The grey dashed line on the plot corresponds to the situation, when actuals and fitted coincide (100% fit). The red line on the plot above is LOESS line, produced by `lowess()` function in R, smoothing the scatterplot to reflect the potential tendencies in the data. In the ideal situation this red line should coinside with the grey line. In addition the variability around the line should not change with the increase of fitted values. In our case there is a slight U-shape in the red line and an insignificant increase in variability around the middle of the data. This could either be due to pure randomness and thus should be ignored, or could indicate a slight non-linearity in the data. After all, we have constructed pure additive model on the data that exhibits seasonality with multiplicative characteristics, which becomes especially apparent at the end of the series, where the drop in level is accompanied by the decrease of variability of the data:

```{r}
plot(adamModelSeat03,7)
```

In order to diagnose this properly, we might use other instruments. One of these is the analysis of standardised residuals. The formula for the standardised residuals will differ depending on the assumed distribution and for some of them comes to the value inside the "$\exp$" part of the [probability density function](#distributions):

1. Normal, $\epsilon_t \sim \mathcal{N}(0, \sigma^2)$: $u_t = \frac{e_t - \bar{e}}{\hat{\sigma}}$;
2. Laplace, $\epsilon_t \sim \mathcal{Laplace}(0, s)$: $u_t = \frac{e_t - \bar{e}}{\hat{s}}$;
3. S, $\epsilon_t \sim \mathcal{S}(0, s)$: $u_t = \frac{e_t - \bar{e}}{\hat{s}^2}$;
4. Generalised Normal, $\epsilon_t \sim \mathcal{GN}(0, s, \beta)$: $u_t = \frac{e_t - \bar{e}}{\hat{s}^{\frac{1}{\beta}}}$;
5. Inverse Gaussian, $1+\epsilon_t \sim \mathcal{IG}(1, s)$: $u_t = \frac{1+e_t}{\bar{e}}$;
6. Gamma, $1+\epsilon_t \sim \mathcal{\Gamma}(s^{-1}, s)$: $u_t = \frac{1+e_t}{\bar{e}}$;
7. Log Normal, $1+\epsilon_t \sim \mathrm{log}\mathcal{N}\left(-\frac{\sigma^2}{2}, \sigma^2\right)$: $u_t = \frac{e_t - \bar{e} +\frac{\hat{\sigma}^2}{2}}{\hat{\sigma}}$.
where $\bar{e}$ is the mean of residuals, which is typically assumed to be zero and $u_t$ is the value of standardised residuals. Note that the scales in the formulae above should be calculated via the formula with the bias correction, i.e. with the division by degrees of freedom, not the number of observations. Also, note that in case of $\mathcal{IG}$, $\Gamma$ and $\mathrm{log}\mathcal{N}$ and additive error models, the formulae for the standardised residuals will be the same, only the assumptions will change (see Section \@ref(ADAMETSAdditiveDistributions)).

Here is an example of a plot of fitted vs standardised residuals in R:
```{r adamModelSeat03Resid, fig.cap="Diagnostics of pure additive ETSX model."}
plot(adamModelSeat03,2)
```

Given that the scale of the original variable is now removed in the standardised residuals, it might be easier to spot the non-linearity. In our case it is still not apparent, but there is a slight U-shape in LOESS line and a slight change in variance. Another plot that we have already used before is standardised residuals over time:

```{r}
plot(adamModelSeat03,8)
```

This plot shows that there is a slight decline in the residuals around year 1977. Still, there is no prominent non-linearity in the residuals, so it is not clear whether any transformations are needed or not.

However, based on my judgment and understanding of the problem, I would expect for the number of injuries and deaths to change proportionally to the change of the level of the data: if after some external interventions the overal level of injuries and deaths would increase, we would expect a percentage decline, not a unit decline with a change of already existing variables in the model. This is why I will try a multiplicative model next:
```{r adamModelSeat05Resid, fig.cap="Diagnostics of pure multiplicative ETSX model."}
adamModelSeat05 <- adam(Seatbelts,"MNM",formula=drivers~PetrolPrice+kms+front+rear+law)
plot(adamModelSeat05,2)
```

The plot shows that the variability is now slightly more uniform across all fitted values, but the difference between Figures \@ref(fig:adamModelSeat03Resid) and \@ref(fig:adamModelSeat05Resid) is not very prominent. One of potential solutions in this situation is to compare the models in terms of information criteria:
```{r}
setNames(c(AICc(adamModelSeat03),AICc(adamModelSeat05)),
         c("Additive model", "Multiplicative model"))
```
Based on this, we would be inclined to select the multiplicative model. My personal judgment in this specific case agrees with the information criterion.


## Model specification: Outliers {#diagnosticsOutliers}
As we discussed in Section \@ref(assumptionsCorrectModel), one of the important assumptions in forecasting and analytics is the correct specification of the model, which also includes "no outliers in the model" element. Outliers might appear for many different reasons:

1. We missed some important information (e.g. promotion) and did not include a respective variable in the model;
2. There was an error in recordings of the data, i.e. a value of 2000 was recorded as 200;
3. We did not miss anything predictable, we just deal with a distribution with fat tails.

In any of these cases, outliers will impact estimates of parameters of our models. In case of ETS, this will lead to higher than needed smoothing parameters, which leads to wider prediction intervals and potentially biased forecasts. In case of ARIMA, the mechanism is more complicated, but also leads to widened intervals and biased forecasts. So, it is important to identify outliers and deal with them.

### Outliers detection
One of the simplest ways for identifying outliers is based on distributional assumptions. For example, if we assume that our data follows normal distribution, then we would expect 95% of observations lie inside the bounds with approximately $\pm 1.96\sigma$ and 99.8% of them to lie inside the $\pm3.09 \sigma$. Sometimes these values are substituted by heuristic "values lying inside 2 / 3 sigmas", which is not precise and works only for Normal distribution. Still, based on this, we could flag the values outside these bounds and investigate them in order to see if any of them are indeed outliers.

Given that ADAM framework supports [different distributions](#distributions), the heuristics mentioned above is not appropriate. We need to get proper quantiles for each of the assumed distributions. Luckily, this is not difficult to do, because the quantile functions for all the distributions supported by ADAM either have analytical forms or can be obtained numerically.

Here is an example in R with the same multiplicative ETSX model and the standardised residuals vs fitted values with the 95% bounds:

```{r}
plot(adamModelSeat05, 2, level=0.95)
```

Note that in case of $\mathcal{IG}$, $\Gamma$ and $\mathrm{log}\mathcal{N}$, the function will plot $\log u_t$ in order to make the plot more readable. The plot demonstrates that there are outliers, some of which are further away from the bounds. Although the amount of outliers is not big, it would make sense investigating why they happened. Well, we know why - we constructed an incorrect model. Given that we deal with time series, plotting residuals vs time is also sometimes helpful:

```{r}
plot(adamModelSeat05, 8)
```

We see that there is no specific pattern in the outliers, they happen randomly, so they appear not because of the omitted variables or wrong transformations. We have 5 observations lying outside the bounds, which given that the sample size of 192 observations, means that the 95% interval contains $\frac{192-9}{192} \times 100 \mathrm{\%} \approx 95.3\mathrm{\%}$ of observations, which is close to the nominal value.

In some cases, the outliers might impact the scale of distribution and will lead to wrong standardised residuals, distorting the picture. This is where studentised residuals come into play. They are calculated similarly to the standardised ones, but the scale of distribution is recalculated for each observation by considering errors on all but the current observation. So, in a general case, this is an iterative procedure which involves looking through $t=\{1,\dots,T\}$ and which should in theory guarantee that the real outliers do not impact the scale of distribution. Here how they can be analysed in R:

```{r}
par(mfcol=c(1,2))
plot(adamModelSeat05, c(3,9))
```

In many cases (ours included) the standardised and studentised residuals will look very similar, but in some cases of extreme outliers they might differ and the latter might show outliers better than the former. 

Given the situation with outliers in our case, we could investigate when they happen in the original data to better understand whether they need to be taken care of. But instead of manually recording, which of the observations lie beyond the bounds, we can get their ids via the `outlierdummy` method from the package `greybox`, which extracts either standardised or studentised residuals and flags those observations that lie outside the constructed interval, automatically creating dummy variables for these observations. Here how it works:
```{r}
adamModelSeat05Outliers <- outlierdummy(adamModelSeat05,
                                        level=0.95, type="rstandard")
```

The method returns several objects (see documentation for details), including the ids of outliers:
```{r}
adamModelSeat05Outliers$id
```

These ids can be used to produce additional plots. For example:

```{r}
plot(actuals(adamModelSeat05))
points(time(Seatbelts[,"drivers"])[adamModelSeat05Outliers$id],
       Seatbelts[adamModelSeat05Outliers$id,"drivers"],
       col="red", pch=16)
text(time(Seatbelts[,"drivers"])[adamModelSeat05Outliers$id],
     Seatbelts[adamModelSeat05Outliers$id,"drivers"],
     adamModelSeat05Outliers$id, col="red", pos=2)
```

Among all these points, there is one special that happens on observation 170. This is when the law for seatbelts was introduced and the model cannot capture the change in injuries and deaths correctly.

::: remark
As a side note, in R, there are several methods for extracting residuals:

- `resid()` or `residuals()` will extract either $e_t$ or $1+e_t$, depending on the distributional assumptions of the model;
- `rstandard()` will extract the standardised residuals $u_t$;
- `rstudent()` will do the same for the studentised ones.

`smooth` package also introduces `rmultistep` which extracts multiple steps ahead in sample forecast errors. We do not discuss this method here, but we might come back to it later in this textbook.
:::

### Dealing with outliers 
Based on the output of `outlierdummy()` method from the previous example, we can construct a model with explanatory variables to interpolate the outliers and neglect their impact on the model:

```{r}
SeatbeltsWithOutliers <- cbind(as.data.frame(Seatbelts[,-c(1,7)]),adamModelSeat05Outliers$outliers)
SeatbeltsWithOutliers$drivers <- ts(SeatbeltsWithOutliers$drivers,
                                    start=start(Seatbelts),
                                    frequency=frequency(Seatbelts))
adamModelSeat06 <- adam(SeatbeltsWithOutliers,"MNM",lags=12,formula=drivers~.)
```

In order to decide, whether the dummy variables help or not, we can use information criteria, comparing the two models:
```{r}
setNames(c(AICc(adamModelSeat05), AICc(adamModelSeat06)),
         c("ETSX","ETSXOutliers"))
```
Comparing the two values above, I would conclude that adding dummies improves the model. But instead of including all of them, we could try the model with the outlier for the suspicious observation 170, which corresponds to the ninth outlier:

```{r}
adamModelSeat07 <- adam(SeatbeltsWithOutliers,"MNM",lags=12,
                        formula=drivers~PetrolPrice+kms+front+rear+law+outlier9)
plot(adamModelSeat07,2)
AICc(adamModelSeat07)
```

This model is slightly worse than both the one with all outliers in terms of AICc, so there are some other dummy variables that improve the fit that might be considered as well, along with the outlier for the observation 170. We could continue the exploration introducing other dummy variables, but in general we should not do that unless we have good reason for that (e.g. we know that something happened that was not captured by the model).


### An automatic mechanism
A similar automated mechanism is implemented in `adam()` function, which has `outliers` parameter, defining what to do with outliers if there are any with the following three options:

1. "ignore" - do nothing;
2. "use" - create the model with explanatory variables as shown in the previous subsection and see if it is better than the simpler model in terms of an information criterion;
3. "select" - create lags and leads of dummies from `outlierdummy()` and then select the dummies based on the [explanatory variables selection mechanism](#ETSXSelection). Lags and leads are needed for cases, when the effect of outlier is carried over to neighbouring observations.

Here how this works for our case:
```{r}
adamModelSeat08 <- adam(Seatbelts,"MNM",lags=12,
                        formula=drivers~PetrolPrice+kms+front+rear+law,
                        outliers="select",level=0.95)
AICc(adamModelSeat08)
```

This automatic procedure will form a matrix that will include original variables together with the outliers, their lags and leads and then select those of them that minimise AICc in a stepwise procedure (discussed in Section \@ref(ETSXSelection)). In our case, the function throws away some of the important variables and sticks with some of outliers. This might also happen because it could not converge to the optimum on each iteration, so increasing `maxeval` might help. Still, given that this is an automated approach, it is prone to potential mistakes and needs to be treated with care as it might select unnecessary dummy variables and lead to overfitting. I would recommend exploring the outliers manually, when possible and not to rely too much on the automated procedures.


### Final remarks
@Koehler2012 explored the question of the impact of outliers on ETS performance in terms of forecasting accuracy. They found that if outliers happen at the end of the time series then it is important to take them into account in a model. If they happen much earlier, then their impact on the final forecast will be negligible. Unfortunately, the authors did not explore the impact of outliers on the prediction intervals, and based on my experience I can tell that the main impact of outliers is on the width of the interval.


## Residuals are i.i.d.: autocorrelation {#diagnosticsResidualsIIDAuto}
One of the typical characteristics of time series models is the dynamic relation between variables. Even if fundamentally the sales of ice cream on Monday do not impact sales of the same ice cream on Tuesday, they might impact advertising expenses or sales of a competing product on Tuesday, Wednesday or next week. Missing this sort of structure might lead to autocorrelation of residuals, which then would impact the estimates of parameters and final forecasts. Autocorrelations might also arise due to wrong transformations of variables, where the model would systematically underforecast the actuals, producing autocorrelated residuals. In this section, we will see one of the potential ways for the regression diagnostics and try to improve the model in a stepwise fashion, trying out different order of [ARIMA](#ADAMARIMA) model.

As an example, we continue with the same seatbelts data, dropping the dynamic part to see what would happen in this case:
```{r}
adamModelSeat09 <- adam(Seatbelts,"NNN",lags=12,
                        formula=drivers~PetrolPrice+kms+front+rear+law)
AICc(adamModelSeat09)
```
There are different ways to diagnose this model. We start with a basic plot of residuals over time:

```{r}
plot(adamModelSeat09,8)
```

We see that on one hand the residuals still contain seasonality and on the other that they do not look stationary. We could conduct ADF and / or KPSS test to get a formal answer to the stationarity question:

```{r}
tseries::kpss.test(resid(adamModelSeat09))
tseries::adf.test(resid(adamModelSeat09))
```
The tests have opposite null hypothesis, and in our case we would reject H$_0$ for both of them on 5% significance level. This means that they contradict each other and we need to use our judgment. First I will see what happens with the model, when we do take differences:

```{r}
# ARIMAX(0,1,0)(0,0,0)_12
adamModelSeat10 <- adam(Seatbelts,"NNN",lags=12,
                        formula=drivers~PetrolPrice+kms+front+rear+law,
                        orders=list(i=1))
AICc(adamModelSeat10)
```
This leads to an improvement in AICc in comparison with the previous model. The residuals of the model are now also much better behaved:
```{r}
plot(adamModelSeat10,8)
```

In order to see whether there are any other dynamic elements left, we will plot [ACF](#ACF) and [PACF](#PACF) of residuals:

```{r}
par(mfcol=c(1,2))
plot(adamModelSeat10,10:11)
```

In case of `adam()` objects, these plots will always have the range for y-axis from -1 to 1 and will start from lag 1 on x-axis. The red horizontal lines represent the "non-rejection" region: if the point lie inside the region, then they are not statistically different from zero on the selected `level` (the uncertainty around them is so high that it covers zero). The points with numbers are those that are statistically significantly different from zero. So, the ACF / PACF analysis might show the lags that are statistically significant on the selected level (the default one is 0.95). Given that this is a statistical instrument, we would expect for approximately (1-level)% (e.g. 5%) of lags lie outside these bounds, so it is fine if we don't see all point lying inside them. However, we should not see any patterns there and we might need to investigate the suspicious lags (low orders of up to 3 - 5 and the seasonal lags if they appear). In our example we see that there is a suspicious lag 1 on ACF and a suspicious lag 2 on the PACF, which might indicate that some dynamic elements are missing (e.g. MA(1) or AR(2)). Furthermore, there are spikes on lag 12 for both ACF and PACF. While it is not clear, what specifically is needed here, we can try out several models and see which one of them is better in order to deremine the appropriate element:
```{r}
# ARIMAX(0,1,0)(1,0,0)_12
adamModelSeat11 <- adam(Seatbelts,"NNN",lags=12,
                        formula=drivers~PetrolPrice+kms+front+rear+law,
                        orders=list(ar=c(0,1),i=1))
AICc(adamModelSeat11)
# ARIMAX(0,1,0)(0,0,1)_12
adamModelSeat12 <- adam(Seatbelts,"NNN",lags=12,
                        formula=drivers~PetrolPrice+kms+front+rear+law,
                        orders=list(i=1,ma=c(0,1)))
AICc(adamModelSeat12)
# ARIMAX(0,1,0)(1,0,1)_12
adamModelSeat13 <- adam(Seatbelts,"NNN",lags=12,
                        formula=drivers~PetrolPrice+kms+front+rear+law,
                        orders=list(ar=c(0,1),i=1,ma=c(0,1)))
AICc(adamModelSeat13)
```
Based on this analysis, we would be inclined to include both seasonal AR(1) and seasonal MA(1) in the model. Next step in our iterative process - another ACF / PACF plot of the residuals:
```{r}
par(mfcol=c(1,2))
plot(adamModelSeat13,10:11)
```

In this case, there is a big spike on ACF for lag 1, so we can try adding MA(1) component in the model:
```{r}
# ARIMAX(0,1,1)(1,0,1)_12
adamModelSeat14 <- adam(Seatbelts,"NNN",lags=12,
                        formula=drivers~PetrolPrice+kms+front+rear+law,
                        orders=list(ar=c(0,1),i=1,ma=c(1,1)))
AICc(adamModelSeat14)
```
Which leads to further improvement in AICc. We could continue our investigations in order to find the most suitable ARIMAX model for the data using this iterative procedure, but this example should suffice in providing the general idea of how it can be done. What we could do else to simplify the process is to use the automated [ARIMA selection algorithm](#ARIMASelection) in `adam()`, which is built on the principles discussed in this section:

```{r}
adamModelSeat15 <- adam(Seatbelts,"NNN",lags=12,
                        formula=drivers~PetrolPrice+kms+front+rear+law,
                        orders=list(ar=c(3,2),i=c(2,1),ma=c(3,2),select=TRUE))
AICc(adamModelSeat15)
```

This new constructed SARIMAX(0,1,1)(0,1,1)$_{12}$ model has lower AICc than the previous one and should be used instead. In fact, it is even better than the `r adamModelSeat05$model` (model 5) from the previous section in terms of AICc, and its residuals are much better behaved than the ones of model 5 (we might need to analyse the residuals for the [potential outliers](#diagnosticsOutliers) in this model though):

```{r}
par(mfcol=c(1,3))
plot(adamModelSeat15,c(8,10:11))
```

So for the purposes of analytics and forecasting, we would be inclined to use SARIMAX(0,1,1)(0,1,1)$_{12}$ rather than `r adamModelSeat05$model`.

As a final word for this section, we have focused our discussion on the visual analysis of time series, ignoring the statistical tests (we only used ADF and KPSS). Yes, there is Durbin-Watson [@WikipediaDurbinWatson2021] test for AR(1) in residuals, and yes there are Ljung-Box [@WikipediaLjungBox2021], Box-Pierce and Breusch–Godfrey [@WikipediaBreuschGodfrey2021] tests for multiple AR elements. But visual inspection of time series is not less powerful than hypothesis testing. In fact, it makes you think and analyse the model and its assumptions, while the tests are the lazy way out that might lead to wrong conclusions because they have the [standard limitations of any hypothesis tests](#hypothesisTesting). After all, if you fail to reject H$_0$ it does not mean that the effect does not exist. Having said that, the statistical tests become extremely useful when you need to process many time series at the same time and cannot physically do visual inspection of them. So, if you are in that situation, I would recommend reading more about them, but I do not aim to retell the content of Wikipedia in this textbook.


## Residuals are i.i.d.: heteroscedasticity {#diagnosticsResidualsIIDHetero}
Another important assumption for conventional models is that the residuals are homoscedastic, meaning that the variance of the residuals stays the same (no matter what). In this section we will see how the issue can be resolved in some cases.
<!-- Typically, this assumption will be violated if the model is not specified correctly. The classical example is the income versus expenditure on meals for different families. If the income is low, then there is not many options what to buy and the variability of expenditures would be low. However, with the increase of the income, the mean expenditures and their variability would increase as well, because there are more options of what to buy, including both cheap and expensive products. If we constructed a basic linear model on such data, then it would violate the assumption of homoscedasticity and as a result will have issues discussed in section \@ref(assumptionsResidualsAreIID). But arguably this would typically appear because of the misspecification of the model. For example, taking logarithms might resolve the issue in many cases, implying that the effect of one variable on the other should be multiplicative rather than the additive. Alternatively, dividing variables by some other variable might (e.g. working with expenses per family member, not per family) resolve the problem as well. Unfortunately, the transformations are not the panacea, so in some cases analyst would need to construct a model, taking the changing variance into account (e.g. GARCH or GAMLSS models). -->

<!-- While in forecasting we are more interested in the holdout performance of models, in econometrics, the parameters of models are typically of the main interest. And, as we discussed [earlier](#assumptionsResidualsAreIID), in case of correctly specified model with heteroscedastic residuals, the estimates of parameters will be unbiased, but inefficient. So, econometricians would use different approaches to diminish the heteroscedasticity effect on parameters: either a different estimator for a model (such as Weighted Least Squares), or a different method for calculation of standard errors of parameters (e.g. Heteroskedasticity-Consistent Standard Errors). This does not resolve the problem, but rather corrects the parameters of the model (i.e. does not heal the illness, but treats the symptoms). Although these approaches typically suffice for the analytical purposes, they do not fix the issues in forecasting. In this section we will see how the issue can be resolved in some cases. -->

### Detecting heteroscedasticity
Building upon our previous example. We will use the ETSX(A,N,A) model, which [as we remember](#diagnosticsTransformations) has some issues. In order to see if the residuals of the model are homoscedastic, we can plot their values against the fitted:

```{r}
par(mfcol=c(1,2))
plot(adamModelSeat03,4:5)
```

These two plots allow detecting specific type of heteroscedasticity, when with the increase of fitted values the residuals variability changes. The plot of absolute residuals vs fitted is more appropriate for models, where the scale parameters is calculated based on absolute values of residuals (e.g. the model with Laplace distribution), while the squared residuals vs fitted shows whether the variance of residuals is stable or not (thus making it more suitable for models with Normal and related distributions). Furthermore, the squared residuals plot might be difficult to read due to outliers, so the first one might help in detecting the heteroscedasticity even, when the scale is supposed to rely on squared errors. What we want to see on these plots, is for all the points to lie in the same corridor for lower and for the higher fitted values and for the red line to be constant. In our case There is a slight increase of the line and the variability of residuals around 1000 is lower than the one around 2000, which might indicate that we have heteroscedasticity in residuals. In our case this is cause by the wrong transformations in the model (see Section \@ref(diagnosticsTransformations)), so the fix of the issue is to use the multiplicative model.

Another diagnostics tool that might become useful in some situations are the plot of absolute and squared standardised residuals versus fitted values. They have a similar idea to the previous plots, but they might change a little because of the standardisation (mean is equal to 0 and scale is equal to 1). These plots become especially useful if the changing variance is modelled explicitly (e.g. via a regression model or a GARCH-type of model. This feature is not yet supported in ADAM):

```{r}
par(mfcol=c(1,2))
plot(adamModelSeat03,13:14)
```

In our case, these plots do not give additional message, we already know that there is a slight heteroscedasticity and that we need to transform the response variable some how (build multiplicative model).

If we suspect that there are some specific variables that might cause heteroscedasticity, we can plot absolute or squared residuals vs these variables to see if they are indeed to blame for that. For example, here how we can produce a basic plot of residuals vs all explanatory variables included in the model:

```{r}
spread(cbind(as.data.frame(abs(resid(adamModelSeat03))),
             adamModelSeat03$data[,all.vars(formula(adamModelSeat03))[-1]]),
       lowess=TRUE)
```

The plot above can be read similarly to the plots discussed above: if we notice a change in variability of residuals or a change (increase of decrease) in the lowess lines with the change of a variable, then this might indicate that the respective variable causes multicollinearity in the model. In our example, it looks like the variable `law` causes the largest issue - all the other variables do not cause as a strong change in variance. We already know that we need to use multiplicative model instead of the additive one in our example, so we will see how the residuals look for the correctly specified model:

```{r}
par(mfcol=c(1,2))
plot(adamModelSeat05,4:5)
```

The plots above do not demonstrate any substantial issues: the residuals look more homoscedastic and given the scale of residuals the change of lowess line does not reflect significant changes in residuals. Additional plot of absolute residuals vs explanatory variables does not show anys serious issues either:

```{r}
spread(cbind(as.data.frame(abs(log(resid(adamModelSeat05)))),
             adamModelSeat05$data[,all.vars(formula(adamModelSeat05))[-1]]),
       lowess=TRUE)
```

So, we can conclude that the multiplicative model resolves the issue with heteroscedasticity. If a variable would still cause an issue with it, it would make sense to construct a model for the variance (e.g. GARCH) in order to address the issue and improve the performance of the model in terms of prediction interval.

There are formal statistical tests for heteroscedasticity, such as White [@WikipediaWhite2021], Breusch-Pagan [@WikipediaBreuschPagan2021] and Bartlett's [@WikipediaBartlett2021] tests. We do not discuss them here for the reason outlined in Section \@ref(diagnosticsResidualsIIDAuto).


## Residuals are i.i.d.: zero expectation
In ADAM framework, this assumption only works for the [additive error models](#ADAMETSPureAdditive). In case of the multiplicative error models, it is changed to "[expectation of the error term is equal to one](#ADAMETSMultiplicativeDistributions)". It does not make sense to check this assumption unconditionally, because it does not mean anything in sample. Yes, it will hold automatically in sample in case of OLS estimation, and the observed mean of the residuals might not be equal to zero in other cases, but this does not give any useful information. In fact, when we work with exponential smoothing models, the in sample residuals being equal to zero might imply for some of them that the final values of components are identical to the initial ones. For example, in case of [ETS(A,N,N)](#SESandETS), we can use the measurement equation from \@ref(eq:ETSANN) to express the final value of level via the previous values up until $t=0$:
\begin{equation}
  \begin{aligned}
    \hat{l}_t &= \hat{l}_{t-1} + \hat{\alpha} e_t = \hat{l}_{t-2} + \hat{\alpha} e_{t-1} + \hat{\alpha} e_t = \\
              & \hat{l}_0 + \hat{\alpha} \sum_{j=1}^t e_{t-j} .
  \end{aligned}
(\#eq:ETSANNMeasurementRecursion)
\end{equation}
If the mean of the residuals in sample is indeed equal to zero then the equation \@ref(eq:ETSANNMeasurementRecursion) reduces to $\hat{l}_t=\hat{l}_0$. So, this assumption cannot be checked in sample, meaning that it is all about the true model and the asymptotic behaviour rather than the model applied to the data.

The only part of this assumption that can be checked is whether the expectation of the residuals *conditional on some variables* is equal to zero (or one). In a way, this comes to making sure that there are no patterns in the residuals and thus no consecutive parts of the data, where residuals have systematically non-zero expectation.

There are different ways to diagnose the issue. The first one is the already discussed plot of standardised (or studentised) residuals vs fitted values from Section \@ref(diagnosticsTransformations). The other one is the plot of residuals over time, something that we have already discussed in Section \@ref(diagnosticsResidualsIIDAuto). In addition, you can also plot residuals vs some of variables in order to see if they cause the change in mean. But in a way all these methods might also mean that the residuals are autocorrelated and / or some transformations of variables are needed.

There is also an effect related to this, which is called "endogeneity". According to econometrics literature it implies that the residuals are correlated with some variables. This becomes equivalent to the situation when the expectation of residuals changes with the change of a variable. The most prominent cause of this is the omitted variables (discussed in Section \@ref(diagnosticsOmitted)), which can be sometimes diagnosed by looking at correlations between the residuals and omitted variables. While econometricians propose using other estimation methods (such as Instrumental Variables) in order to diminish the effect of endogeneity, the forecasters cannot allow themselves doing that, because we need to fix the problem in order to get more adequate forecasts. Unfortunately, there is no universal recipe for the solution of this problem, but in some cases transforming variables, adding the omitted ones or substituting them by [proxies](https://en.wikipedia.org/wiki/Proxy_(statistics)) (if we the variables are unavailable) might resolve the problem.

<!-- The assumption for the h steps ahead? -->


## Residuals are i.i.d.: distributional assumptions
Finally, we come to the distributional assumptions of ADAM. As discussed earlier (for example, in Section \@ref(ADAMETSEstimationLikelihood)), ADAM framework supports several distributions, and the specific parts of assumptions will change depending on the type of the error term in the model. Given that, it is relatively straightforward to see if the residuals of the model follow the assumed distribution or not. There exist several tools for that.

The simplest one is called Quantile-Quantile (QQ) plot. It produces a figure with theoretical vs actual quantiles and shows, whether they are close to each other or not. Here is, for example, how the QQ plot will look for one of the previous models, assuming Normal distribution:

```{r}
plot(adamModelSeat03,6)
```

If the residuals do not contradict the assumed distribution, then all the points should lie either very close to or on the line. In our case, the majority of points are close to the line, but the tails are slightly off. In ADAM, this might mean that we should either use a different error type or a different distribution. Just for the sake of argument, we can try ETSX(M,N,M) model, with the same set of explanatory variables as in the model `adamModelSeat03`, and with the same Normal distribution:

```{r}
adamModelSeat16 <- adam(Seatbelts,"MNM",formula=drivers~PetrolPrice+kms+front+rear+law,
                        distribution="dnorm")
plot(adamModelSeat16,6)
```

According to the new QQ plot, the residuals of the new model are much closer to the theoretical ones, there is now only the right tail that is wrong - the actual values are a bit further away than expected. This can be addressed by using a skewed distribution, for example, Inverse Gaussian:

```{r}
adamModelSeat17 <- adam(Seatbelts,"MNM",formula=drivers~PetrolPrice+kms+front+rear+law,
                        distribution="dinvgauss")
plot(adamModelSeat17,6)
```

The new QQ plot demonstrates that the empirical residuals follow the assumed distribution much closer than in the previous cases: there are just few observations that lie slightly away from the line, but they could happen at random. So, based on this simple analysis we could conclude that Inverse Gaussian distribution is more suitable for this situation than the Normal one. Interestingly, this is supported by the AIC values, which very roughly reflect the same thing:

```{r}
setNames(c(AIC(adamModelSeat03),AIC(adamModelSeat16),AIC(adamModelSeat17)),
         c("Additive Normal","Multiplicative Normal","Multiplicative IG"))
```

Another way to analyse the distribution of residuals is to plot histogram together with the theoretical density function. Here is an example:

```{r}
hist(residuals(adamModelSeat03), probability=TRUE)
lines(seq(-250,250,1),
      dnorm(seq(-250,250,1), 0, sd(residuals(adamModelSeat03))),
      col="red")
```

However, this plot is much more difficult to analyse than QQ plot, because of the bars, which average out the quantiles. So, in general I would recommend using QQ plots instead.

There are also formal tests for the distribution of residuals, such as Shapiro-Wilk [@WikipediaShapiroWilk2021] and Kolmogorov-Smirnov [@WikipediaKolmogorovSmirnov2021]. The former tests the hypothesis that residuals follow Normal distribution, while the latter one is much more flexible and allows comparing the empirical distribution with any other (theoretical or empirical). However, I prefer to use visual inspection, when possible instead of doing these tests because, as we discussed earlier in Section \@ref(hypothesisTesting), the null hypothesis is always wrong, and it will inevitably be rejected with the increase of the sample size. Besides, if you fail to reject H$_0$, it does not mean that your variable follows the assumed distribution, it only means that you have not found enough evidence to reject it.


## Multicollinearity
One of the classical issues in econometrics and in statistics in regression context is the issue of multicollinearity. In a way, this has nothing to do with classical assumptions of linear regression, because it is unreasonable to assume that the explanatory variables have some specific relation between them - they are what they are, and multicollinearity mainly causes issues with estimation of the parameters of model, not with its structure. But it is an issue nonetheless, so it is worth exploring.

Multicollinearity appears, when either some of explanatory variables are correlated with each other (see Section \@ref(correlationCoefficient)), or their linear combination explains another explanatory variable included in the model. Depending on the strength of this relation and the estimation method used for model construction, the multicollinearity might cause issues of varying severity. For example, in the case, when two variables are perfectly correlated (correlation coefficient is equal to 1 or -1), the model will have perfect multicollinearity and it would not be possible to estimate its parameters. Another example is a case, when an explanatory variable can be perfectly explained by a set of other explanatory variables (resulting in $R^2$ being close to one), which will cause exactly the same issue. The classical example of this situation is the dummy variables trap, when all values of categorical variable are used in regression together with the constant resulting in the linear relation $\sum_{j=1}^k d_j = 1$. Given that the square root of $R^2$ of linear regression is equal to multiple correlation coefficient, these two situations are equivalent and just come to "absolute value of correlation coefficient is equal to 1". Finally, as discussed briefly in Section \@ref(assumptionsXreg), if correlation coefficient is high, but not equal to one, the effect of multicollinearity will lead to [less efficient estimates](#estimatesPropertiesEfficiency) of parameters. The loss of efficiency is in this case proportional to the absolute value of correlation coefficient. All of this tells us how this problem can be diagnosed and that this diagnoses should be carried out before constructing regression model.

First, we can calculate correlation matrix for the available variables. If they are all numeric, then `cor()` function from `stats` should do the trick (we remove the response variable from consideration):

```{r}
cor(Seatbelts[,-2])
```

This matrix tells us that there are some variables that are highly correlated and might reduce efficiency of estimates of parameters of regression model if included in the model together. This applies to `DriversKilled` and `front` and also to `front` and `rear`. However, the values are not very high, so this might mean that there won't be a serious issue in the model. If we have a mix of numerical and categorical variables, then `assoc()` (aka `association()`) function from `greybox` will help. It will detect the categorical variables automatically and will select between correlation and other appropriate measures of association (such as multiple correlation and Cramer's V). In our case, all the variables are numeric, so it would produce a similar matrix. If you want to see this function in action, see what it produces for the `mtcars` data from `datasets`:

```{r eval=FALSE}
assoc(mtcars)
```

In order to cover the second situation with linear combination of variables, we can use the `determ()` (aka `determination()`) function from `greybox`:

```{r}
determ(Seatbelts[,-2])
```

This function will construct linear regression models for each variable from all the other variables and report the $R^2$ from these models. If there are coefficients of determination close to one, then this might indicate that the variables would cause multicollinearity in the model. In our case, we see that `front` is linearly related to other variables, and we can expect it to cause the reduction of efficiency of estimate of parameters. If we remove the `DriversKilled` from consideration (we do not want to include it in our model anyway), then the picture will change:

```{r}
determ(Seatbelts[,-c(1,2)])
```

In this case, the effect multicollinearity will be lower than in the previous situation, but it will still impact the estimates of parameters, making them less efficient than in the ideal situation.

Instead of calculating the coefficients of determination, econometricians prefer to calculate Variance Inflation Factor (VIF), which shows by how many times the estimates of parameters will loose efficiency. Its formula is based on the $R^2$ calculated above:
\begin{equation*}
  \mathrm{VIF}_j = \frac{1}{1-R_j^2}
\end{equation*}
for each model $j$. Which in our case can be calculated as:
```{r}
1/(1-determ(Seatbelts[,-c(1,2)]))
```
This is useful when you want to see the specific impact on the variance of parameters, but is difficult to work with, when it comes to model diagnostics, because the value of VIF lies between zero and infinity. So, I prefer using the determination coefficients instead, which is always bounded by $(0, 1)$ region and thus easier to interpret.

Furthermore, as you might have noticed, the discussion here was mainly focused on regression models. When it comes to dynamic models, then the situation might differ.

In case of conventional [ARIMA model](#ARIMA), multicollinearity is inevitable by construct, because of the autocorrelations between actual values. This is why sometimes Heteroskedasticity- and autocorrelation-consistent (HAC) estimators of the covariance matrix [see Section 15.4 of @Hanck2020] of parameters are used instead of the standard ones. They are designed to fix the issue and produce standard errors of parameters that are close to those without the issue.

Finally, in case of state space models, and specifically in case of [ETS](#ADAMX), the multicollinearity might not cause as serious issues as in the case of regression. For example, it is possible to use all the values of [categorical variable](#ETSXDynamicCategories), avoiding the trap of dummy variables. The values of categorical variable in this case are considered as changes relative to the baseline. The classical example of this is the seasonal model, for example, ETS(A,A,A), where the seasonal components can be considered as a set of parameters for dummy variables, expanded from the seasonal categorical variable (e.g. months of year variable). If we set $\gamma=0$, thus making the seasonality deterministic, the ETS can still be estimated even though all values of the variable are used. This becomes apparent with the conventional ETS model, for example, from `forecast` package for R:

```{r}
etsModel <- forecast::ets(AirPassengers, "AAA")
# Calculate determination coefficients for seasonal states
determ(etsModel$states[,-c(1:2)])
```

As we see, the states of the model are almost perfectly correlated, but still the model works and does not have issue that the classical linear regression would have.

<!-- ## Model diagnostics on high frequency data -->
<!-- ## Model diagnostics for intermittent model -->
