# Forecasting with ADAM {#ADAMForecasting}
Finally, we come to the technicalities of producing forecasts using ADAM. We have already discussed how conditional expectations can be generated from some of the models (e.g. Sections \@ref(pureAdditiveExpectationAndVariance), \@ref(pureMultiplicativeExpectationAndVariance), \@ref(ADAMARIMARecursiveMoments), \@ref(ADAMXConventionalConditionalMoments) and \@ref(ADAMXDynamicMoments)), but we have not discussed this in necessary detail. Furthermore, as discussed in Section \@ref(forecastingPlanningAnalytics), forecasts should align with specific decisions, but we have not discussed how to do that.

In this chapter, we start with an explanation of how the simulation paths can be generated to obtain moments and quantiles in the cases when they are not available analytically (Section \@ref(ADAMForecastingSimulations)). We then discuss the principles behind calculating the conditional moments from ADAM (including ETS, ARIMA, regression and their combinations, Section \@ref(ADAMForecastingMoments)). We then discuss various methods for prediction interval construction, starting from the basic parametric and ending with empirical ones and those that take the uncertainty of parameters into account (building upon Chapter \@ref(ADAMUncertainty)). Finally, we discuss prediction intervals for the intermittent state space model (Chapter \@ref(ADAMIntermittent)), one-sided intervals and cumulative forecasts over the forecast horizon, which is useful in practice, especially when inventory decisions need to be made. We also discuss the confidence interval for the conditional mean, which is not as important as the other topics mentioned above but is still worth discussing.


## Creating simulation paths {#ADAMForecastingSimulations}
As mentioned earlier in previous chapters, for some models, the conditional h steps ahead moments do not have analytical expressions. For example, as discussed in Section \@ref(pureMultiplicativeExpectationAndVariance), pure multiplicative models typically do not have formulae for the conditional expectations and variance for longer horizons. The one exception is the ETS(M,N,N) model, where the point forecast corresponds to the conditional expectation for any horizon as long as the expectation of the error term $1+\epsilon_t$ is equal to one. This problem is not only important for the moments, but it also arises when quantiles of distributions are needed, especially for models with multiplicative error. In all these cases, when the moments and/or quantiles are not available analytically, they need to be obtained via other means. One of those is simulating trajectories and then calculating moments numerically.

### Simulating trajectories {#ADAMForecastingExpectationSimulations}
The general idea of this approach is to use the estimated parameters, the last obtained state vector (level, trend, seasonal, ARIMA components etc.) and the estimate of the scale of distribution to generate the possible paths of the data. The simulation itself is done in several steps after obtaining parameters, states and the scale:

1. Generate $h$ random variables for the error term, $\epsilon_{t+j}$ or $1+\epsilon_{t+j}$ -- depending on the type of error, assumed distribution in the model (the latter was discussed in Sections \@ref(ADAMETSAdditiveDistributions) and \@ref(ADAMETSMultiplicativeDistributions)) and estimated parameters of distribution (typically, scale);
2. Insert the error terms in the state space model \@ref(eq:ETSADAMStateSpace) from Section \@ref(ADAMETSGeneral), both in the transition and observation parts, do that iteratively from $j=1$ to $j=h$:
\begin{equation*}
  \begin{aligned}
    {y}_{t+j} = &w(\mathbf{v}_{t+j-\mathbf{l}}) + r(\mathbf{v}_{t+j-\mathbf{l}}) \epsilon_{t+j} \\
    \mathbf{v}_{t+j} = &f(\mathbf{v}_{t+j-\mathbf{l}}) + g(\mathbf{v}_{t+j-\mathbf{l}}) \epsilon_{t+j}
  \end{aligned},
\end{equation*}
3. Record actual values for $j=\{1, \dots, h\}$;
4. Repeat (1) - (3) $n$ times;
5. Take desired moments or quantiles for each horizon from 1 to $h$.

Graphically, the result of this approach is shown in Figure \@ref(fig:adamScenarios), where each separate path is shown in grey colour, and the expectation is in black. See the R code for this in Subsection \@ref(scenariosExample).

```{r adamScenarios, fig.cap="Sample paths (scenarios) generated from ADAM for the holdout sample.", echo=FALSE}
adam(BJsales, "YYN", h=10, holdout=TRUE) |>
    forecast(h=10, interval="simulated", scenarios=TRUE, nsim=500) ->
    adamForecast
plot(as.vector(adamForecast$mean), ylim=range(adamForecast$scenarios), type="l",
     xlab="Horizon", ylab="Sample paths")
for(i in 1:ncol(adamForecast$scenarios)){
    lines(adamForecast$scenarios[,i], col=rgb(0.8,0.8,0.8,0.4))
}
lines(as.vector(adamForecast$mean), lwd=2)
```

::: remark
In the case of multiplicative trend or multiplicative seasonality, it makes sense to take trimmed mean instead of the basic arithmetic one on step (5). This is because the models with these components might exhibit explosive behaviour, and thus the expectation might become unrealistic. I suggest using 1% trimming, although this does not have any scientific merit and is only based on my personal expertise.
:::

The simulation-based approach is universal, no matter what model is used, and can be applied to any ETS, ARIMA, regression model or combination (including dynamic ETSX, intermittent demand and multiple frequency models). Furthermore, instead of extracting moments on step 5, one can take geometric mean, median or any desired statistics.

The main issue with this approach is that the conditional expectation or any other statistics calculated based on this will differ with every new simulation run. If $n$ is small, these values will be less stable (vary more with the new runs). But they will reach some asymptotic values with the increase of $n$, staying random nonetheless. However, this is a good thing because this randomness reflects the uncertain nature of these statistics in the sample: the true values are never known, and the estimates will inevitably change with the sample size change. Another limitation is the computational time and memory usage: the more iterations we want to produce, the more calculations will need to be done, and more memory will be used. Luckily, time complexity in this situation should be linear: $O(h \times n)$.

### Demonstration in R {#scenariosExample}
In order to demonstrate the simulation approach, we consider an artificial case of ETS(M,M,N) model with $l_t=1000$, $b_t=0.95$, $\alpha=0.1$, $\beta=0.01$, and Gamma distribution for error term with scale $s=0.05$. We generate 1000 scenarios from this model for the horizon of $h=10$ using `sim.es()` function from `smooth` package:
```{r}
nsim <- 1000
h <- 10
s <- 0.1
initial <- c(1000,0.95)
persistence <- c(0.1,0.01)
y <- sim.es("MMN", obs=h, nsim=nsim, persistence=persistence,
            initial=initial, randomizer="rgamma",
            shape=1/s, scale=s)
```
After running the code above, we will obtain an object `y` that will contain several variables, including `y$data` with all the 1000 possible future trajectories. We can plot them to get an impression of what we are dealing with (see Figure \@ref(fig:adamForecastSimulated)):

```{r adamForecastSimulated, fig.cap="Data generated from 1000 ETS(M,M,N) models."}
plot(y$data[,1], ylab="Sales", ylim=range(y$data),
     col=rgb(0.8,0.8,0.8,0.4), xlab="Horizon")
# Plot all the generated lines
for(i in 2:nsim){
  lines(y$data[,i], col=rgb(0.8,0.8,0.8,0.4))
}
# Add conditional mean and quantiles
lines(apply(y$data,1,mean))
lines(apply(y$data,1,quantile,0.025),
      col="grey", lwd=2, lty=2)
lines(apply(y$data,1,quantile,0.975),
      col="grey", lwd=2, lty=2)
```

Based on the plot in Figure \@ref(fig:adamForecastSimulated), we can see what the conditional h steps ahead expectation (black line) and what the 95% prediction interval will be for the data based on the ETS(M,M,N) model with the parameters mentioned above.

Similar paths are produced by `forecast()` function for `adam` class in `smooth`. If you need to extract them for further analysis, they are returned in the object `scenarios` if the parameters `interval="simulated"` and `scenarios=TRUE` are set.


## Conditional moments and scale {#ADAMForecastingMoments}
We have already discussed how to obtain conditional expectation and variance in Sections \@ref(pureAdditiveExpectationAndVariance) and \@ref(pureMultiplicativeExpectationAndVariance). However, the topic is worth discussing in more detail, especially for non-normal distributions.

### Conditional expectation {#ADAMForecastingExpectation}
The general rule that applies to ADAM in terms of generating conditional expectations is that if you deal with the pure additive model, then you can produce forecasts analytically. This not only applies to ETS but also to ARIMA (Subsection \@ref(ADAMARIMARecursiveMoments)) and regression (Section \@ref(ADAMXConventionalConditionalMoments)). If the model has multiplicative components (such as multiplicative error or trend or seasonality) or is formulated in logarithms (as, for example, ARIMA in logarithms), then simulations (Section \@ref(ADAMForecastingSimulations)) should be preferred - the point forecasts from these models would not necessarily correspond to the conditional expectations.

### Explanatory variables
If the model contains explanatory variables, then the h steps ahead conditional expectations should use them in the calculation. The main challenge in this situation is that future values might not be known in some cases. This has been discussed in Section \@ref(ADAMXConventionalConditionalMoments). Practically speaking, if the user provides the holdout sample values of explanatory variables, the `forecast.adam()` method will use them in forecasting. If they are not provided, the function will produce forecasts for each of the explanatory variables via the `adam()` function and use the conditional h steps ahead expectations in forecasting.

### Conditional variance and scale {#ADAMForecastingVariance}
Similar to conditional expectations, as we have discussed in Sections \@ref(pureAdditiveExpectationAndVariance) and \@ref(pureMultiplicativeExpectationAndVariance), the conditional h steps ahead variance is in general available only for the pure additive models. While the conditional expectation might be required on its own to use as a point forecast, the conditional variance is typically needed to produce prediction intervals. However, it becomes useful only in cases of distributions that support convolution (addition of random variables), which limits its usefulness to pure additive models and to additive models applied to the data in logarithms. For example, if we deal with Inverse Gaussian distribution, then the h-steps-ahead values will not follow Inverse Gaussian distribution, and we would need to revert to simulations in order to obtain the proper statistics for it. Another situation would be a multiplicative error model that relies on Normal distribution - the product of Normal distributions is not a Normal distribution, so the statistics would need to be obtained using simulations again.

If we deal with pure additive model with either Normal, Laplace, S or Generalised Normal distributions, then the formulae derived in Section \@ref(pureAdditiveExpectationAndVariance) can be used to produce h-steps-ahead conditional variance. Having obtained those values, we can then produce conditional h-steps-ahead scales for the distributions (which would be needed, for example, to generate quantiles from these distributions), using the relations between the variance and scale in those distributions (discussed in Section \@ref(ADAMETSAdditiveDistributions)):

1. Normal: scale is $\sigma^2_h$;
2. Laplace: $s_h = \sigma_h \sqrt{\frac{1}{2}}$;
3. S: $s_h = \sqrt{\sigma_h}\sqrt[4]{\frac{1}{120}}$;
4. Generalised Normal: $s_h = \sigma_h \sqrt{\frac{\Gamma(1/\beta)}{\Gamma(3/\beta)}}$.

If the variance is needed for the other combinations of model/distributions, simulations would need to be done to produce multiple trajectories, similar to how it was done in Section \@ref(ADAMForecastingSimulations). An alternative to this would be the calculation of in-sample multistep forecast errors (similar to how it was discussed in Sections \@ref(multistepLosses) and \@ref(diagnosticsResidualsIIDExpectationMultiple)). Having the matrix of forecast errors, we can then calculate variance for each horizon $h$.


### Scale model
In the case of the scale model (Chapter \@ref(ADAMscaleModel)), the situation becomes more complicated because we no longer assume that the variance of the error term is constant (homoscedastic) -- we now assume that it is a model on its own. In this case, we need to take a step back to the recursion \@ref(eq:ETSADAMStateSpacePureAdditiveRecursion04) and when taking the variance, introduce the time-varying variance $\sigma_{t+h}^2$.

::: remark
Note the difference between $\sigma_{t+h}^2$ and $\sigma_{h}^2$ in our notations - the former is the variance of the error term for the specific step $t+h$, while the latter is the conditional variance $h$ steps ahead, which is derived based on the assumption of homoscedasticity.
:::

Making that substitution leads to the following analytical formula for the h-steps-ahead conditional variance in the case of the scale model:
\begin{equation}
    \text{V}(y_{t+h}|t) = \sum_{i=1}^d \left(\mathbf{w}_{m_i}^\prime \sum_{j=1}^{\lceil\frac{h}{m_i}\rceil-1} \mathbf{F}_{m_i}^{j-1} \mathbf{g}_{m_i} \mathbf{g}^\prime_{m_i} (\mathbf{F}_{m_i}^\prime)^{j-1} \mathbf{w}_{m_i} \sigma_{t+h-j}^2 \right) + \sigma_{t+h}^2 .
    (\#eq:ETSADAMFullWithScaleModelVariance)
\end{equation}
This variance can then be used, for example, to produce quantiles from the assumed distribution.

As mentioned above, in the case of the not purely additive model or model with other distributions than Normal, Laplace, S or Generalised Normal, the conditional variance can be obtained using simulations. In the case of the scale model, the principles will be the same, just assuming that each error term $\epsilon_{t+h}$ has its own scale, obtained from the estimated scale model. The rest of the logic will be exactly the same as discussed in Section \@ref(ADAMForecastingSimulations).


## Prediction intervals {#ADAMForecastingPI}
As discussed in Section 6.5 of @SvetunkovSBA, a prediction interval is needed to reflect the uncertainty about the data. In theory, the 95% prediction interval will cover the actual values in 95% of the cases if the model is correctly specified. The specific formula for prediction interval will vary with the assumed distribution. For example, for the Normal distribution (assuming that $y_{t+j} \sim \mathcal{N}(\mu_{y, t+j}, \sigma_j^2)$) we will have the classical one:
\begin{equation}
    y_{t+j} \in (\hat{y}_{t+j} + z_{\alpha/2} \hat{\sigma}_j, \hat{y}_{t+j} + z_{1-\alpha/2} \hat{\sigma}_j),
    (\#eq:predictionInterval)
\end{equation}
where $\hat{y}_{t+j}$ is the estimate of $\mu_{y,t+j}$, $j$ steps ahead conditional expectation, $\hat{\sigma}_j^2$ is the estimate of ${\sigma}_j^2$, $j$ steps ahead variance of the error term obtained using principles discussed in Subsection \@ref(ADAMForecastingVariance) (for example, calculated via the formula \@ref(eq:ETSADAMStateSpacePureAdditiveRecursionMeanAndVariance)) and $z$ is z-statistics (quantile of standard normal distribution) for the selected significance level $\alpha$.

::: remark
Note that $\alpha$ has nothing to do with the smoothing parameters for the level of ETS models.
:::

This type of prediction interval can be called **parametric**. It assumes a specific distribution and relies on the other assumptions about the constructed model (such as residuals are i.i.d., see Section \@ref(diagnostics)). Most importantly, it assumes that the $j$ steps ahead value follows a specific distribution related to the one for the error term. In case of normal distribution, the assumption $\epsilon_t\sim\mathcal{N}(0,\sigma^2)$ implies that $y_{t+1}\sim\mathcal{N}(\mu_{y,t+1}, \sigma_1^2)$ and due to *convolution* of random variables (the sum of random variables follows the same distribution as individual variables, but with different parameters), that $y_{t+j}\sim\mathcal{N}(\mu_{y,t+j},\sigma_j^2)$ for all $j$ from 1 to $h$.

The interval produced via \@ref(eq:predictionInterval) corresponds to two quantiles from the normal distribution and can be written in a more general form as:
\begin{equation}
    y_{t+j} \in \left(q \left(\hat{y}_{t+j},\hat{\sigma}_j^2,\frac{\alpha}{2}\right), q\left(\hat{y}_{t+j},\hat{\sigma}_j^2,1-\frac{\alpha}{2}\right)\right),
    (\#eq:predictionIntervalGeneral)
\end{equation}
where $q(\cdot)$ is a quantile function of an assumed distribution, $\hat{y}_{t+j}$ acts as a location and $\hat{\sigma}^2$ acts as a scale of distribution. Using this general formula \@ref(eq:predictionIntervalGeneral) for prediction interval, we can construct them for other distributions as long as they support convolution. In the ADAM framework, this works for all pure additive models that have error term that follows one of the following distributions:

1. Normal: $\epsilon_t \sim \mathcal{N}(0, \sigma^2)$, thus $y_{t+j} \sim \mathcal{N}(\mu_{y, t+j}, \sigma_j^2)$;
2. Laplace: $\epsilon_t \sim \mathcal{Laplace}(0, s)$ and $y_{t+j} \sim \mathcal{Laplace}(\mu_{y, t+j}, s_j)$;
3. S: $\epsilon_t \sim \mathcal{S}(0, s)$, $y_{t+j} \sim \mathcal{S}(\mu_{y, t+j}, s_j)$;
4. Generalised Normal: $\epsilon_t \sim \mathcal{GN}(0, s, \beta)$, so that $y_{t+j} \sim \mathcal{GN}(\mu_{y, t+j}, s_j, \beta)$.

If a model has multiplicative components or relies on a different distribution, then the several steps ahead actual value will not necessarily follow the assumed distribution and the formula \@ref(eq:predictionIntervalGeneral) will produce incorrect intervals. For example, if we work with a pure multiplicative ETS model, ETS(M,N,N), assuming that $\epsilon_t \sim \mathcal{N}(0, \sigma^2)$, the two-steps-ahead actual value will be expressed in terms of the values on observation $t$:
\begin{equation}
    y_{t+2} = l_{t+1} (1+\epsilon_{t+2}) = l_{t} (1+\alpha \epsilon_{t+1}) (1+\epsilon_{t+2}) ,
    (\#eq:ETSMNN2Steps)
\end{equation}
which introduces the product of normal distributions, and thus $y_{t+2}$ does not follow Normal distribution anymore. In such cases, we might have several options of what to use to produce intervals. They are discussed in the subsections below.


### Approximate intervals
Even if the actual multistep value does not follow the assumed distribution, we can use approximations in some cases: the produced prediction interval will not be too far from the correct one. The main idea behind the approximate intervals is to rely on the same distribution for $y_{t+j}$ as for the error term, even though we know that the variable will not follow it. In the case of multiplicative error models, the limit \@ref(eq:limitOf1x) can be used to motivate the usage of that assumption. For example, in the case of the ETS(M,N,N) model, we know that $y_{t+2}$ will not follow the normal distribution, but if the variance of the error term is low (e.g. $\sigma^2 < 0.05$) and the smoothing parameter $\alpha$ is close to zero, then the normal distribution would be a satisfactory approximation of the real one. This becomes clear if we expand the brackets in \@ref(eq:ETSMNN2Steps):
\begin{equation}
    y_{t+2} = l_{t} (1 + \alpha \epsilon_{t+1} + \epsilon_{t+2} + \alpha \epsilon_{t+1} \epsilon_{t+2}) .
    (\#eq:ETSMNN2Steps2)
\end{equation}
With the conditions discussed above (low $\alpha$, low variance) the term $\alpha \epsilon_{t+1} \epsilon_{t+2}$ will be close to zero, thus making the sum of normal distributions dominate in the formula \@ref(eq:ETSMNN2Steps2). The advantage of this approach is in its speed: you only need to know the scale parameter of the error term and the conditional expectation. The disadvantage of the approach is that it becomes inaccurate with the increase of parameters and scale of the model. The rule of thumb when to use this approach: if the smoothing parameters are all below 0.1 (in the case of ARIMA, this is equivalent to MA terms being negative and AR terms being close to zero) or the scale of distribution is below 0.05, the differences between the proper interval and the approximate one should be negligible.


### Simulated intervals {#ADAMForecastingPISimulations}
This approach relies on the idea discussed in Section \@ref(ADAMForecastingSimulations). It is universal and supports any distribution because it only assumes that the error term follows the selected distribution (no need for the actual value to do that as well).

The simulated paths are then produced based on the generated values and the assumed model. After generating $n$ paths, one can take the desired quantiles to get the bounds of the interval. The main issue of the approach is that it is time-consuming (slower than the approximate intervals) and might be highly inaccurate if the number of iterations $n$ is low. This approach is used as a default in `adam()` for the non-additive models.


### Semiparametric intervals
The three approaches above assume that the residuals of the applied model are i.i.d. (see discussion in Chapter \@ref(diagnostics)). If this assumption is violated (for example, the residuals are autocorrelated), then the intervals might be miscalibrated (i.e. producing wrong values). In this case, we might need to use different approaches. One of these is the construction of **semiparametric** prediction intervals [see, for example, @Lee2014]. This approach relies on the in-sample multistep forecast errors discussed in Subsection \@ref(diagnosticsResidualsIIDExpectationMultiple). After producing $e_{t+j|t}$ for all in sample values of $t$ and for $j=1,\dots,h$, we can use these errors to calculate the respective h steps ahead conditional variances $\sigma_j^2$ for $j=1,\dots,h$. These values can then be inserted in the formula \@ref(eq:predictionIntervalGeneral) to get the desired prediction intervals. The approach works well in the case of pure additive models, as it relies on specific assumed distribution. However, it might have limitations similar to those discussed earlier for the mixed models and the models with positively defined distributions (such as Log-Normal, Gamma and Inverse Gaussian). It can be considered a semiparametric alternative to the approximate method discussed above.


### Nonparametric intervals
When some of the assumptions might be violated, and when we cannot rely on the parametric distributions, we can use the **nonparametric** approach proposed by @Taylor1999. The authors proposed using the multistep forecast errors to construct the following quantile regression model:
\begin{equation}
    \hat{e}_{t+j} = a_0 + a_1 j + a_2 j^2,
    (\#eq:IntervalsQuantileReg)
\end{equation}
for each of the bounds of the interval. The motivation behind the polynomial in \@ref(eq:IntervalsQuantileReg) is because, typically, the multistep conditional variance will involve the square of the forecast horizon. The main issue with this approach is that the polynomial function has an extremum, which might appear sometime in the future. For example, the upper bound of the interval would increase until that point and then start declining. To overcome this limitation, I propose using the power function instead:
\begin{equation}
    \hat{e}_{t+j} = a_0 j^a_1 .
    (\#eq:IntervalsQuantileRegIS)
\end{equation}
This way, the bounds will always change monotonically, and the parameter $a_1$ will control the speed of expansion of the interval. The model \@ref(eq:IntervalsQuantileRegIS) is estimated using quantile regression for the upper and the lower bounds separately, as @Taylor1999 recommend. This approach does not require any assumptions about the model and works as long as there are enough observations in-sample (so that the matrix of forecast errors contains more rows than columns). The main limitation of this approach is that it relies on quantile regression and thus will have the same issues as, for example, pinball score has (see discussion in Section \@ref(uncertainty)): the quantiles are not always uniquely defined. Another limitation is that we assume that the quantiles will follow the model \@ref(eq:IntervalsQuantileRegIS), which might be violated in real-life scenarios.


### Empirical intervals {#ADAMForecastingPIEmpirical}
Another alternative to the parametric intervals uses the same matrix of multistep forecast errors, as discussed earlier. The **empirical** approach is more straightforward than the approaches discussed above and does not rely on any assumptions [it was discussed in @Lee2014]. The idea behind it is just to take quantiles of the forecast errors for each forecast horizon $j=1,\dots,h$. These quantiles are then added to the point forecast if the error term is additive or are multiplied by it in the case of the multiplicative one. @Trapero2019 show that the empirical prediction intervals perform on average better than the analytical ones. So, in general, I would recommend producing empirical intervals if it was not for the computational difficulties related to the multistep forecast errors. If you have an additive model and believe that the assumptions are satisfied, then the parametric interval will be as accurate but faster. Furthermore, the approach will be unreliable on small samples due to the same problem with the quantiles discussed earlier.


### Complete parametric intervals
So far, **all the intervals discussed above** relied on an unrealistic assumption that the parameters of the model are known. This is one of the reasons why the intervals produced for ARIMA and ETS are typically narrower than expected [see, for example, results of tourism competition, @Athanasopoulos2011]. But as we discussed in Section \@ref(ADAMUncertainty), there are ways of capturing the uncertainty of estimated parameters of the model and propagating it to the future uncertainty (i.e. to conditional h steps ahead variance). As discussed in Section \@ref(adamRefitted), the more general approach is to create many in-sample model paths based on randomly generated parameters of the model. This way, we can obtain a variety of states for the final in-sample observation $T$ and then use those values to construct final prediction intervals. The simplest and most general way of producing intervals, in this case, is using simulations (as discussed earlier in Subsection \@ref(ADAMForecastingPISimulations)). The intervals produced via this approach will be wider than the conventional ones, and their width will be proportional to the uncertainty around the parameters. This also means that the intervals might become too wide if the uncertainty is not captured correctly (see discussion in Section \@ref(adamRefitted)). One of the main limitations of the approach is its computational time: it will be proportional to the number of simulation paths for both refitted model and prediction intervals.

It is also theoretically possible to use other approaches for the intervals construction in case of complete uncertainty (e.g. "empirical" one for each of the set of parameters of the model), but they would be even more computationally expensive than the approach described above and will have the limitations similar to the discussed above (i.e. non-uniqueness of quantiles, sample size requirements).


### Explanatory variables
In all the cases described above, when constructing prediction intervals for the model with explanatory variables, we assume that their values are known in the future. Even if they are not provided by the user and need to be forecasted, the produced conditional expectations will be used for all the calculations. This is not an entirely correct approach, as was shown in Subsection \@ref(ADAMXConventionalConditionalMomentsRandom), but it speeds up the calculation process and typically produces adequate results.

The more theoretically correct approach is to take the multistep variance of explanatory variables into account. This would work for pure additive models for explanatory and response variables but imply more complicated formulae for other models.


### Example in R {#ADAMForecastingPIExample}
All the types of intervals discussed in this Section are implemented for the `adam()` models in `smooth` package. In order to demonstrate how they work and how they differ, we consider the example with ETS model on `BJSales` data:
```{r}
adamETSBJ <- adam(BJsales, h=10, holdout=TRUE)
modelType(adamETSBJ)
```
The model selected above is ETS(`r substr(modelType(adamETSBJ),1,1)`,`r substr(modelType(adamETSBJ),2,nchar(modelType(adamETSBJ))-1)`,`r substr(modelType(adamETSBJ),nchar(modelType(adamETSBJ)),nchar(modelType(adamETSBJ)))`). In order to make sure that the parametric intervals are suitable, we can do model diagnostics (see Section \@ref(diagnostics)), producing plots shown in Figure \@ref(fig:adamModelDiagnostics).

```{r adamModelDiagnostics, fig.cap="Diagnostics of the ADAM on BJSales data."}
par(mfcol=c(2,3))
plot(adamETSBJ,c(2,4,6,8,10:11))
```

The model's residuals do not exhibit any serious issues. Given that this is a pure additive model, we can conclude that the parametric intervals would be appropriate for this situation and produce the plot with them.

The only thing that these intervals do not take into account is the uncertainty about the parameters, so we can construct the respective intervals either via the `reforecast()` function or using the same `forecast()`, but with the option `interval="complete"`. Note that this is a computationally expensive operation (both in terms of time and memory), so the more iterations you set up, the longer it will take and the more memory it will consume. The two intervals are shown next to each other in Figure \@ref(fig:adamModelPIComplete):

```{r adamModelPIComplete, fig.cap="Prediction intervals for ADAM on BJSales data."}
par(mfcol=c(1,2))
plot(forecast(adamETSBJ,h=10,interval="parametric"),
     main="Parametric prediction interval", ylim=c(200,280))
plot(forecast(adamETSBJ,h=10,interval="complete",nsim=100),
     main="Complete prediction interval", ylim=c(200,280))
```

The resulting complete parametric interval shown in Figure \@ref(fig:adamModelPIComplete) will not be substantially different from the parametric one, maybe slightly wider. This is because the smoothing parameters of the model are high, thus the model forgets the initial states fast, and the uncertainty does not propagate to the last observation as much as in the case of lower values of parameters:
```{r}
summary(adamETSBJ)
```

Figure \@ref(fig:adamModelPIRefitted) demonstrates what happens with the fitted values when we take the uncertainty into account.

```{r adamModelPIRefitted, fig.cap="Refitted values for ADAM on BJSales data."}
plot(reapply(adamETSBJ))
```

As we see from Figure \@ref(fig:adamModelPIRefitted), the uncertainty around the line is narrow at the end of the sample, so the prediction intervals are similar between the two methods.

To make things more complicated and exciting, we introduce explanatory variable with lags and leads of the indicator `BJsales.lead`, automatically selecting the model and explanatory variables using information criteria (see discussion in Chapter \@ref(ADAMSelection)).

```{r}
BJsalesData <- cbind(as.data.frame(BJsales),
                     xregExpander(BJsales.lead,c(-3:3)))
colnames(BJsalesData)[1] <- "y"
adamETSXBJ <- adam(BJsalesData, "YYY",
                   h=10, holdout=TRUE,
                   regressors="select")
```

I have asked the function specifically to do the selection between pure multiplicative models (see Section \@ref(ETSSelection)). We will then construct several types of prediction intervals and compare them:

```{r warning=FALSE}
adamETSXBJPI <- vector("list",6)
intervalType <- c("approximate", "semiparametric",
                  "nonparametric", "simulated",
                  "empirical", "complete")
for(i in 1:6){
  adamETSXBJPI[[i]] <- forecast(adamETSXBJ, h=10,
                                interval=intervalType[i])
}
```

These can be plotted in Figure \@ref(fig:adamModelXPI6Plots).

```{r adamModelXPI6Plots, fig.cap="Different prediction intervals for ADAM ETS(M,N,N) on BJSales data"}
par(mfcol=c(3,2), mar=c(2,2,2,1))
for(i in 1:6){
  plot(adamETSXBJPI[[i]],main=paste0(intervalType[i]," interval"))
}
```

The thing to notice is how the width and shape of intervals change depending on the used approach. The approximate and simulated intervals look very similar because the selected model is ETSX(M,N,N) with a standard error of `r round(sigma(adamETSXBJ),3)` (thus, the approximation works well). The complete interval is similar because the estimated smoothing parameter $\alpha=1$ (hence, the forgetting happens instantaneously). However, it has a slightly different shape because the number of iterations for the interval was low (`nsim=100` for `interval="complete"` by default). The semiparametric interval is the widest as it calculates the forecast errors directly but still uses the normal approximation. Both nonparametric and empirical are skewed because the in-sample forecast errors followed skewed distributions, which can be seen via the plot (see Figure \@ref(fig:adamModelXForecastErrors)):

```{r adamModelXForecastErrors, fig.cap="Distribution of in-sample multistep forecast errors from ADAM ETSX(M,N,N) model on BJSales data. Red point correspond to mean values."}
adamETSXBJForecastErrors <- rmultistep(adamETSXBJ,h=10)
boxplot(1+adamETSXBJForecastErrors)
abline(h=1,col="red")
points(apply(1+adamETSXBJForecastErrors,2,mean),
       col="red",pch=16)
```

Analysing the plot \@ref(fig:adamModelXPI6Plots), it might be challenging to select the most appropriate type of prediction interval. But the model diagnostics (Section \@ref(diagnostics)) might help in this situation:

1. If the residuals look i.i.d. and the model does not omit important variables, then choose between parametric, approximate and simulated intervals, choose:
a. "parametric" in case of the pure additive model,
b. "approximate" in other cases, when the standard error is lower than 0.05 or smoothing parameters are close to zero,
c. "simulated" if you deal with a non-additive model with high values of standard error and smoothing parameters,
d. "complete parametric" when the smoothing parameters of the model are close to zero, and you want to take the uncertainty about the parameters into account;
2. If residuals seem to follow the assumed distribution but are not i.i.d., then the "semiparametric" approach might help. Note that this only works on samples of $T>>h$;
3. If residuals do not follow the assumed distribution, but your sample is still longer than the forecast horizon, then use either "empirical" or "nonparametric" intervals.

Finally, `forecast.adam()` will automatically select between "parametric", "approximate" and "simulated" if you ask for `interval="prediction"`.


## Other aspects of forecast uncertainty {#forecastingADAMOther}
There are other elements related to forecasting and taking uncertainty into account that we have not discussed in the previous sections. Here we discuss several special cases where forecasting approaches might differ from the conventional ones.

### Constructing intervals for intermittent demand model
When it comes to constructing prediction intervals for the intermittent state space model (see Section \@ref(ADAMIntermittent)), then there is an important aspect that should be taken into account. Given that the model consists of two parts: demand sizes and demand occurrence, the prediction interval should take the uncertainty from both parts into account. In this case, we should first predict the probability of occurrence of demand for the h steps ahead and then decide what the width of the interval should be based on this probability. For example, if we estimate that the demand will occur with probability $\hat{p}_{t+h|t} = 0.8$, then this means that we expect that in 20% of the cases, we will observe zeroes. This should reduce the confidence level for the demand sizes. Formally speaking, this comes to the following equation:
\begin{equation}
    F_y(y_{t+h} \leq q) = \hat{p}_{t+h|t} F_z(z_{t+h} \leq q) +(1 -\hat{p}_{t+h|t}),
    (\#eq:statCDF)
\end{equation}
where $F_y(\cdot)$ is the cumulative distribution function of demand, $F_z(\cdot)$ is the cumulative distribution function of the demand sizes, $\hat{p}_{t+h|t}$ is the h steps ahead expected probability of occurrence and $q$ is the quantile of distribution. In the formula \@ref(eq:statCDF), we know the expected probability and we know the confidence level $F_y(y_{t+h} \leq q)$. The unknown element is the $1-\alpha = F_z(z_{t+h} \leq q)$. So after regrouping elements we get:
\begin{equation}
    F_z(z_{t+h} \leq q) = \frac{F_y(y_{t+h} \leq q) -(1 -\hat{p}_{t+h|t})}{\hat{p}_{t+h|t}}.
    (\#eq:statCDFDemandSizes)
\end{equation}
For example, if the confidence level is 0.95 and the expected probability of occurrence is 0.8, then $F_z(z_{t+h} \leq q) = \frac{0.95 -0.2}{0.8} = 0.9375$. Assuming a distribution for the demand sizes, we can use formula \@ref(eq:predictionIntervalGeneral) to construct the parametric prediction interval. Alternatively, we can use any other approach discussed in Section \@ref(ADAMForecastingPI) to generate the intervals.


### One-sided prediction interval {#forecastingADAMOtherOneSided}
In some cases, we might not need both bounds of the interval. For example, when we deal with intermittent demand, we know that the lower bound will be equal to zero in many cases. In fact, this will always happen when the significance level is lower than the probability of inoccurrence $1-p_{t+h|t}$: we will have the quantile equal to zero because the probability of having zeroes is higher than the significance level. Another example is the safety stock calculation: we only need the upper bound of the interval, and we need to make sure that the specific proportion of demand is satisfied (e.g. 95% of it). In these cases, we can just focus on the particular bound of the interval and drop the other one. Statistically speaking, this means that we cut only one tail of the assumed distribution. This has its implications and issues in several scenarios:

- When we are interested in **upper bound** only and deal with **positive distribution** of demand (for example, Gamma, Log-Normal or Inverse Gaussian), we know that the demand will always lie between zero and the constructed bound. In cases of low volume (or even intermittent) data, this makes sense because the original data might contain zeroes or have values close to it. The upper bound in this case will be lower than in the case of the two-sided prediction interval because we would not be splitting the probability into two parts (for the left and the right tails);
- The combination of **lower bound** and **positive distribution** implies that the demand will be greater than the specified bound in the pre-selected number of cases (defined by confidence level). There is no natural bound from above, so from a theoretical point of view, this implies that the demand can be infinite;
- **Upper** or **lower** bound with **real-valued distribution** (such as Normal, Laplace, S or Generalised Normal) implies that the demand is either below or above the specified level, respectively, without any natural limit on the other side. If Normal distribution is used on positive low volume data, there is a natural lower bound, but the model itself will not be aware of it and will not restrict the space with the specific value.


### Cumulative over the horizon forecast {#forecastingADAMOtherCumulative}
Another related thing to consider when producing forecasts in practice is that the point forecast is not needed in some contexts. Instead, the cumulative over the forecast horizon (or over the lead time) might be more suitable. The classic example is the safety stock calculation based on the lead time (time between the order of product and its delivery). In this situation, we need to make sure that while the product is being delivered, we do not run out of stock, thus still satisfying the selected level of demand (e.g. 95%), but now over the whole period of time rather than on every separate observation.

In the case of **pure additive ADAM**, there are analytical formulae for the conditional expectations and conditional variance for this case that can be used in forecasting. These formulae come directly from the recursive relation \@ref(eq:ETSADAMStateSpacePureAdditiveRecursion04) (for derivations in a simpler case, see for example, @Hyndman2008b and @Svetunkov2017):
```{r echo=FALSE}
    knitr::asis_output('\\begin{equation}')
if(knitr:::is_latex_output()){
    knitr::asis_output('\\resizebox{\\textwidth}{!}{$')
}
```
    \begin{aligned}
        \mu_{Y,t,h} = \text{E}(Y_{c,t,h}|t) = & \sum_{j=1}^h \sum_{i=1}^d \left(\mathbf{w}_{m_i}'     \mathbf{F}_{m_i}^{\lceil\frac{j}{m_i}\rceil-1} \right) \mathbf{v}_{t} \\
        \sigma^2_{Y,h} = \text{V}(Y_{c,t,h}|t) = & \left(1 + \sum_{k=1}^{h-1} \left(1+ (h-k) \sum_{i=1}^d \left(\mathbf{w}_{m_i}' \sum_{j=1}^{\lceil\frac{k}{m_i}\rceil-1} \mathbf{F}_{m_i}^{j-1} \mathbf{g}_{m_i} \mathbf{g}'_{m_i} (\mathbf{F}_{m_i}')^{j-1} \mathbf{w}_{m_i} \right) \right) \right) \sigma^2
    \end{aligned},
```{r echo=FALSE}
if(knitr:::is_latex_output()){
    knitr::asis_output('$}')
}
    knitr::asis_output('(\\#eq:ETSADAMStateSpacePureAdditiveRecursionMeanAndVarianceCumulative)')
    knitr::asis_output('\\end{equation}')
```
where $Y_{c,t,h}=\sum_{j=1}^h y_{t+j}$ is the cumulative actual value and all the other variables have been defined in Section \@ref(adamETSPureAdditiveRecursive). Based on these expectation and variance, we can construct prediction interval as discussed in Section \@ref(ADAMForecastingPI).

In cases of **multiplicative and mixed ADAM**, there are no closed forms for the conditional expectation and variance. As a result, simulations similar to the one discussed in Section \@ref(ADAMForecastingSimulations) are needed to produce all possible paths for the next $h$ steps ahead. The main difference would be that before taking the expectation or quantiles, the paths would need to be aggregated over the forecast horizon $h$. This approach, together with the idea of a one-sided prediction interval, can be directly used to calculate the safety stock over the lead time.


### Example in R {#forecastingADAMOtherExample}
For the demonstration purposes, we consider an artificial intermittent demand example, similar to the one from Section \@ref(IntermittentExample):
```{r}
y <- ts(c(rpois(20,0.25), rpois(20,0.5), rpois(20,1),
          rpois(20,2), rpois(20,3)))
```
For simplicity, we apply iETS(M,Md,N) model with odds ratio occurrence:

```{r}
adamiETSy <- adam(y, "MMdN", occurrence="odds-ratio",
                  h=7, holdout=TRUE)
plot(adamiETSy,7)
```

To make this setting closer to a possible real-life situation, we assume that the lead time is seven days, and we need to satisfy the 99% of demand for the last seven observations based on our model. Thus we produce the upper bound for the cumulative values for the confidence level of 99%:
```{r}
adamiETSyForecast <- forecast(adamiETSy, h=7,
                              cumulative=TRUE,
                              interval="prediction",
                              side="upper")
```
Given that we deal with cumulative values, the basic plot will not be helpful, we should produce a different one (see Figure \@ref(fig:adamModeliETSCumulative)):

```{r adamModeliETSCumulative, fig.cap="The actual cumulative demand (black dot), the expectation (the solid blue line) and the 95% quantile of the distribution of the cumulative demand (the dashed grey line) based on iETS model."}
plot(sum(adamiETSy$holdout), ylab="Cumulative demand",
     xlab="", xaxt="n", pch=16,
     ylim=range(c(0, sum(adamiETSy$holdout),
                  adamiETSyForecast$upper)))
abline(h=adamiETSyForecast$mean, col="blue",
       lwd=2)
abline(h=adamiETSyForecast$upper, col="grey",
       lwd=2, lty=2)
```

What Figure \@ref(fig:adamModeliETSCumulative) demonstrates is that for the holdout period of 7 days, the cumulative demand was around `r sum(adamiETSy$holdout)` units, while the upper bound of the interval was approximately `r round(adamiETSyForecast$upper,0)`. Based on that upper bound, we could place an order (based on what we already have in stock) and have an appropriate safety stock.

To see if the approach is suitable, we would need to apply it in either a rolling origin fashion (Section \@ref(rollingOrigin)) or to a set of products to collect the distribution of related error measures.


### Confidence interval
Finally, we can construct a confidence interval for some statistics. As discussed in Section 6.4 of @SvetunkovSBA, it can be built for the mean, a parameter, fitted values, etc. In our context, we might be interested in the confidence interval for the conditional h steps ahead expectation. This implies that we are interested in the uncertainty of the line, not of the actual values, which can only be constructed for the model that takes the uncertainty of parameters into account (as discussed in Section \@ref(ADAMUncertainty)). The construction of confidence interval, in this case, relies on the normal distribution, based on Central Limit Theorem, as long as the basic assumptions for the model and CLT are satisfied [see Section 6.2 and Chapter 15 of @SvetunkovSBA]. Technically speaking, the construction of confidence interval comes to capturing the model uncertainty discussed in Chapter \@ref(ADAMUncertainty).

#### Example in R
The only way how the confidence interval can be constructed for ADAM is via the `reforecast()` function. Consider the example with ADAM ETS(A,Ad,N) on `BJSales` data as in Section \@ref(ADAMForecastingPIExample):
```{r}
adamETSBJ <- adam(BJsales, h=10, holdout=TRUE)
```
The confidence interval for this model can be produced either directly via `reforecast()` or via `forecast()`, which will call it for you:

```{r eval=FALSE}
adamETSBJConfidence <- forecast(adamETSBJ, h=10,
                                interval="confidence", nsim=1000)
plot(adamETSBJConfidence)
```

Note that I have increased the number of iterations for the simulation to get a more accurate confidence interval around the conditional expectation. This will consume more memory, as the operation involves creating 1000 sample paths for the fitted values and another 1000 for the holdout sample forecasts.

```{r adamModelConfidence, fig.cap="Confidence interval for the point forecast from ADAM ETS(A,Ad,N) model.", echo=FALSE}
load("data/adamModelConfidence.Rdata")
plot(adamModelConfidence)
```

Figure \@ref(fig:adamModelConfidence) shows the uncertainty around the point forecast based on the uncertainty of parameters of the model. As can be seen, the interval is narrow, demonstrating that the conditional expectation would not change much if the model's parameters would vary slightly. The fact that the actual values are systematically above the forecast does not mean anything because the confidence interval does not consider the uncertainty of actual values. 
