# Case studies {#CaseStudies}
In this chapter, we consider case studies, demonstrating how ADAM can be used for specific tasks. At the moment, we only have one case study: number of visitors of the [openforecast.org](https://www.openforecast.org) blog.

## Openforecast.org blog visits
For this example, we consider statistics of the number of visitors of the [openforecast.org](https://www.openforecast.org) blog. This statistics does not include the visits to the online monographs published on the website and only focuses on the Wordpress blog, which has existed since 2015. It started as a blog with materials written in Russian but then slowly but surely has started focusing on the international audience. The [statistics was collected](http://top.mail.ru/jump?from=2629777) by a Russian service "Mail.RU" via a counter at the bottom of the website. Although the audience has changed over the years, the recent (10th October 2024) data tells that the largest group of visitors (18.38%) still comes from Russia, which is mainly because I have several pages on basics of forecasting in Russian language. I downloaded the data in csv files, aggregated it to daily level, and as a result got a 3502 observations of my website visitors.

A thing to note is that whenever I publish posts, I usually advertise them on LinkedIn and they are also reposted automatically on [R-bloggers](https://www.r-bloggers.com/). This boosts the number of visitors, but typically does not change the core audience.

As I usually say, forecasting and analytics need to be done to inform decisions. If they are done just for the sake of doing them, this might be a waste of time. So, what decision can I make based on this statistics? I decided that the main decision I want to make is to decide, when to publish a new post to get a larger audience. I also want to understand the impact of LinkedIn/R-bloggers reposts on the number of visitors. So, I formulated the following questions:

1. When should I publish my new post to get a higher coverage (larger audience of readers).
2. What is the average effect of reposts on LinkedIn on the number of website visitors?
3. What is the average effect of reposts on R-bloggers on the number of website visitors?

To answer these questions, I decided to use ADAM, compare it with several simple models to decide what works better and what can give me a better answer.

### The data

```{r echo=FALSE}
load("data/openForecastVisits.Rdata")
```

The data is available [on github of the monograph in the Rdata format](https://github.com/config-i1/adam/blob/master/data/openForecastVisits.Rdata). It is a `zoo` object of daily data with specific dates.

```{r}
plot(openForecastVisits)
```
