# adam() in R Cheat Sheet {#cheatSheet}
This Chapter summarises the main ways of using `adam()` in R with the references to the chapters and sections, where specific elements are discussed.

Estimate the best ETS model for the provided data (Section \@ref(ETSSelection)):
```{r eval=FALSE}
ourModel <- adam(y)
```

The best ETS with a holdout of 10 observations to test the performance of model:
```{r eval=FALSE}
ourModel <- adam(y, h=10, holdout=TRUE)
```

Build a pure multiplicative ETS with seasonality and an arbitrary seasonal lag of 7 (can be applied to an object of any class, Section \@ref(ETSSelection)):
```{r eval=FALSE}
ourModel <- adam(y, model="YYM", lags=7)
```

Estimate the best ARIMA model for the provided data (assuming seasonal lag of 7, Section \@ref(ARIMASelection)):
```{r eval=FALSE}
ourModel <- adam(y, model="NNN", lags=7,
                 orders=list(ar=c(3,2),i=c(2,1),ma=c(3,2),select=TRUE))
```

Build ARIMA(0,1,1) with drift (Chapter \@ref(ARIMA) for the general ARIMA and Section \@ref(ARMAConstant) for the one with constant):
```{r eval=FALSE}
ourModel <- adam(y, model="NNN", lags=7,
                 orders=c(0,1,1), constant=TRUE)
```

Estimate ETS(A,N,N)+ARIMA(1,0,0) model:
```{r eval=FALSE}
ourModel <- adam(y, model="ANN", orders=c(1,0,0))
```

Use Generalised Normal distribution for the residuals of ADAM ETS(A,A,N) (Sections \@ref(ADAMETSAdditiveDistributions), \@ref(ADAMETSMultiplicativeDistributions) and \@ref(ADAMETSEstimationLikelihood)):
```{r eval=FALSE}
ourModel <- adam(y, model="AAN", distribution="dgnorm")
```

Select the best distribution for the specific ADAM ETS(A,A,N) (Chapter \@ref(ADAMSelection)):
```{r eval=FALSE}
ourModel <- auto.adam(y, model="AAN")
```

Select the most appropriate ETSX model for the provided data (which can be any 2-dimensional object, Chapter \@ref(ADAMX)):
```{r eval=FALSE}
ourModel <- adam(data)
```

Specify, which explanatory variables to include and in what form (Section \@ref(ADAMXFormulation)):
```{r eval=FALSE}
ourModel <- adam(data, formula=y~x1+x2+I(x2^2))
```

Select the set of explanatory variables for ETSX(M,N,N) based on AIC (Section \@ref(ETSXSelection)):
```{r eval=FALSE}
ourModel <- adam(data, model="MNN",
                 regressors="select", ic="AIC")
```

Estimate ETS(A,Ad,N) model using a multistep loss function, GTMSE (Section \@ref(multistepLosses)):
```{r eval=FALSE}
ourModel <- adam(y, model="AAdN",
                 h=10, loss="GTMSE")
```

Estimate ARIMA(1,1,2) using a multistep loss function (Section \@ref(multistepLosses)) with backcasting of initials:
```{r eval=FALSE}
ourModel <- adam(y, model="NNN", orders=c(1,1,2),
                 h=10, loss="GTMSE", initial="backcasting")
```

Select and estimate the most appropriate ETS model on the data with multiple frequencies (Chapter \@ref(ADAMMultipleFrequencies)):
```{r eval=FALSE}
ourModel <- adam(y, model="ZXZ", lags=c(24,24*7,24*365))
```

Select and estimate the triple seasonal ARIMA on the data with multiple frequencies (Chapter \@ref(ADAMMultipleFrequencies)):
```{r eval=FALSE}
ourModel <- adam(y, model="NNN", lags=c(24,24*7,24*365),
                 orders=list(ar=c(3,2,2,2),i=c(2,1,1,1),ma=c(3,2,2,2),select=TRUE),
                 initial="backcasting")
```

Apply an automatically selected the occurrence part of the model to intermittent data (Section \@ref(ADAMOccurrence)):
```{r eval=FALSE}
oesModel <- oes(y, model="YYY", occurrence="auto")
```

Use the estimated occurrence model in `adam()` to model intermittent data (Section \@ref(ADAMDemandSizes)):
```{r eval=FALSE}
ourModel <- adam(y, model="YYY", occurrence=oesModel)
```

Or alternatively just use the same model for occurrence and demand sizes part (Chapter \@ref(ADAMIntermittent)):
```{r eval=FALSE}
ourModel <- adam(y, model="YYY", occurrence="auto")
```

Estimate the scale model for previously estimated ADAM (Chapter \@ref(ADAMscaleModel)):
```{r eval=FALSE}
scaleModel <- sm(ourModel, model="YYY")
```

Implant the scale model into the ADAM for future use (e.g. for forecasting):
```{r eval=FALSE}
finalModel <- implant(ourModel, scaleModel)
```

Produce diagnostics plots to see if the ADAM can be improved any further (Chapter \@ref(diagnostics)):
```{r eval=FALSE}
par(mfcol=c(2,2))
plot(ourModel, which=c(1,2,4,6))
```

Extract conventional, standardised and studentised residuals (Chapter \@ref(diagnostics)):
```{r eval=FALSE}
residuals(ourModel)
rstandard(ourModel)
rstudent(ourModel)
```

Plot time series decomposition according to ADAM ETS (Section \@ref(ETSTaxonomy)):
```{r eval=FALSE}
plot(ourModel, which=12)
```

Produce point forecast and prediction interval from ADAM for 10 steps ahead (Chapter \@ref(ADAMForecasting)):
```{r eval=FALSE}
forecast(ourModel, h=10, interval="prediction")
```

Produce point forecast and prediction interval for ADAM, cumulative over the lead time of 10 (Subsection \@ref(forecastingADAMOtherCumulative)):
```{r eval=FALSE}
forecast(ourModel, h=10, interval="prediction",
         cumulative=TRUE)
```

Produce point forecast and empirical prediction interval for upper bound (upper quantile of distribution, Sections \@ref(ADAMForecastingPIEmpirical) and \@ref(forecastingADAMOtherOneSided)):
```{r eval=FALSE}
forecast(ourModel, h=10, interval="empirical",
         side="upper")
```

Produce summary of ADAM (Chapter \@ref(ADAMUncertainty)):
```{r eval=FALSE}
summary(ourModel)
```

Reapply ADAM with randomly selected initials and parameters and produce forecasts from each of these models (Section \@ref(adamRefitted)):
```{r eval=FALSE}
reapply(ourModel)
reforecast(ourModel)
```

Extract multistep forecast errors from ADAM (Subsection \@ref(diagnosticsResidualsIIDExpectationMultiple)):
```{r eval=FALSE}
rmultistep(ourModel, h=10)
```

Extract covariance matrix of multistep forecast errors from ADAM (Section \@ref(multistepLosses)):
```{r eval=FALSE}
multicov(ourModel, h=10)
```

Extract actual values and fitted from ADAM:
```{r eval=FALSE}
actuals(ourModel)
fitted(ourModel)
```
